---
title: "Rmarkdown_16S_Ananlysis"
author: "Simon Vandersanden"
date: "2026-03-05"
output: html_document
---

Loading in the necessary libraries for basic handeling 
```{r}
library(phyloseq)
library(dplyr)
library(picante)    # For Phylogenetic Diversity
library(microbiome) # For Dominance metrics
```


## 1:Inspecting the different parts of the phyloseq object
```{r}
#Load in the phyloseq object
source("scripts/01_load_phyloseq.R") 
# To see the OTU table
mibi_otu_table<- as(otu_table(mibi_phylo_object), "matrix")
head(mibi_otu_table[,1:5])
# To see the Taxonomy
mibi_tax_table <- as(tax_table(mibi_phylo_object), "matrix")
head(mibi_tax_table)
# To see the Metadata
mibi_meta_table <- as(sample_data(mibi_phylo_object), "data.frame")
head(mibi_tax_table)

# You can also swap the metadata table, in case you want to make adjustments to the grouping of the samples or add some new metadata the the specific samples
mibi_meta_table_new <- read.table("data/META_Griftpark_Enrichments_Form.txt", header = T, sep = "\t", row.names = 1, stringsAsFactors = FALSE, dec = ',')

sample_data(mibi_phylo_object) <- mibi_meta_table_new

#After this step you have the otu count table information, the taxonomy table and the correct or an updated version of the metadata

```
## 2.1:Data Pre-processing: Filtering
```{r}
#Normally some pre filtering has been done on the reads before they were put into the phyloseq object: the sequencing depth threshhold, the decontamination, and some taxonomic filtering of certain undesired taxa such as Eukaryota, Mitochondria and Chloroplast sequences.

# It is possible that that some blancs and mock samples are still present in the samples, these should be filtered out because we will only filter using all the 'real samples'
#filter out these blanc samples:
samples_to_remove <- c("Blank", "MockB", "nosample1", "nosample2", "nosample3", "nosample4")
mibi_phylo_object <- prune_samples(
  !(sample_names(mibi_phylo_object) %in% samples_to_remove),
  mibi_phylo_object
)
# Check the rarefaction curves:
source("scripts/02_rarefaction_curves.R")

# --- Filtering parameters:USER INPUT SECTION ---
# Thresholds
prev_filter_threshold = 1   # Present in at least X samples
abund_filter_threshold = 5 # At least X reads 
# Logic Switches (1 = On, 0 = Off)
  # Here you can specify which filtering must be done,
  # Only based on prevalence, or abudance. Then fill in a 0 accordingly
  # Most of the time it is a combination of both prevalence and abundance, for example, at least >10 reads in 5 samples.
prev_filter = 1 
abund_filter = 1
# --------------------------
# Run the background script
source("scripts/02_filtering.R")
# The resulting object 'mibi_phylo_object_filt' is now ready for further processing

#Subset selection, this is optional, you can select a specific set of samples.
# Specify the group you want to isolate
target_group <- "2A"
# Filter the phyloseq object to keep ONLY group 2A
mibi_phylo_object_filt <- subset_samples(mibi_phylo_object, group == target_group)
# IMPORTANT: Remove ASVs that are no longer present in this specific group
# (Otherwise you have 1346 ASVs but many might have 0 counts in group 2A)
mibi_phylo_object_filt <- prune_taxa(taxa_sums(mibi_phylo_object_filt) > 0, mibi_phylo_object_filt)

```
2.2:Data Pre-processing: Normalisation
```{r}
## 2.2 Normalization & The "Zero Problem"
#Microbiome data is **compositional**. Because the total number of reads is an artifact of the DNA sequencer, the counts are not absolute; they are "parts of a whole." If one taxon blooms, others appear to decrease simply because they occupy a smaller percentage of the total pool.

### Handling Sparsity
#Our filtered matrix still contains many zeros. To perform the **CLR transformation**, we must address these because $\log(0)$ is undefined. We can use a **Pseudo-count (+1)** to shift the data, or more advanced methods such as Bayesian-Multiplicative Replacement. https://www.rdocumentation.org/packages/zCompositions/versions/1.6.0/topics/cmultRepl 

### Transformation Applied
We generate two versions of the data:
1.  **TSS (Relative Abundance)**: Used for stacked bar plots and descriptive statistics.
2.  **CLR (Centered Log-Ratio)**: Used for all downstream statistical modeling (PCoA, Correlation, Differential Abundance).

```{r normalization}
# Source the normalization script
source("../scripts/02_normalization.R")

# Quick check of the first sample (first 5 ASVs) in CLR space
head(abundances(mibi_clr)[1, 1:5])


```

3.1: Calculating alfa diversity metrics
```{r}
#Alpha diversity metrics are a general term for metrics that describe the species richness, evenness, or diversity within a sample. Collectively, these metrics contribute to a comprehensive set of traits characterizing the samples, allowing for the determination of key aspects of the microbial community.
#4 main categories exist: richness, dominance, phylogenetic and information metrics

alfa_diversity_phyloseq_object <-mibi_phylo_object_filt
# --- USER INPUT SECTION: Select Community Traits ---
# 1. Richness (How many?)
selected_richness <- c('observed', 'chao1') 

# 2. Dominance (Who is the power player?)
selected_dominance <- c('absolute', 'dbp') 

# 3. Evenness (How fair is the distribution?)
selected_evenness <- c('pielou', 'simpson', 'evar')

# 4. Information (Mathematical entropy/uncertainty)
selected_information <- c('shannon') 
# --------------------------------------------------
# Run background script
source("scripts/04_alpha_diversity.R")
# Display results
knitr::kable(head(alpha_output))

```
3.1: Alpha diversity visualisation
```{r}

#We visualize the distribution of community traits across our experimental groups. Note that the Y-axis scales are "free" because the units for Richness (counts) and Dominance (0-1) are fundamentally different.  
  
# --- PLOT SETTINGS: USER INPUT ---
plot_x_axis <- "group"     # What is on the X-axis? (e.g., "Timepoint", "group")
plot_color_by <- "group"       # What defines the colors? (e.g., "group", "replicate")
plot_metrics <- all_requested_cols # Use all calculated, or specific ones: c("observed", "diversity_shannon")

# Optional: Wrap the plot into how many columns?
plot_cols <- 3 

# ---------------------------------
source("scripts/04_1_alpha_plots.R")


plot(alpha_boxplot)


```






