---
title: "Rmarkdown_16S_Ananlysis_mibibio"
author: "Simon Vandersanden"
date: "2026-03-05"
output: html_document
---

Loading in the necessary libraries for basic handeling 
```{r}
library(phyloseq)
library(dplyr)
library(picante)    # For Phylogenetic Diversity
library(microbiome) # For Dominance metrics
library(vegan)      # For Rarefaction curves
library(zCompositions)
library(ggplot2)
library(ggrepel)
library(MicrobiomeStat)
```
## 1:Inspecting the different parts of the phyloseq object
```{r,include=TRUE, results='hide'}
#Load in the phyloseq object
mibi_phylo_object <- readRDS("data/ps_IdTax_noncontam_NoMitChlo_Tree.rds") 
# To see the OTU table
mibi_otu_table<- as(otu_table(mibi_phylo_object), "matrix")
#head(mibi_otu_table[,1:5])
# To see the Taxonomy
mibi_tax_table <- as(tax_table(mibi_phylo_object), "matrix")
#head(mibi_tax_table)
# To see the Metadata
mibi_meta_table <- as(sample_data(mibi_phylo_object), "data.frame")
#head(mibi_tax_table)

# You can also swap the metadata table, in case you want to make adjustments to the grouping of the samples or add some new metadata the the specific samples
mibi_meta_table_new <- read.table("data/META_Griftpark_Soilcore_Measurements.txt", header = T, sep = "\t", row.names = 1, stringsAsFactors = FALSE, dec = ',')
#Make sure the rownames of the loaded in metadata match the sample names in the phyloseq object
stopifnot(all(sample_names(mibi_phylo_object) %in% rownames(mibi_meta_table_new)))

sample_data(mibi_phylo_object) <- mibi_meta_table_new

#After this step you have the otu count table information, the taxonomy table and the correct or an updated version of the metadata

```
## 2.1:Data Pre-processing: Filtering out all non-samples and checking rarefaction curves
```{r,include=TRUE}
#Normally some pre filtering has been done on the reads before they were put into the phyloseq object: the sequencing depth threshhold, the decontamination, and some taxonomic filtering of certain undesired taxa such as Eukaryota, Mitochondria and Chloroplast sequences.

# It is possible that that some blancs and mock samples are still present in the samples, these should be filtered out because we will only filter using all the 'real samples'
#filter out these blanc samples:
samples_to_remove <- c("Blank", "MockB", "nosample1", "nosample2", "nosample3", "nosample4")
mibi_phylo_object <- prune_samples(
  !(sample_names(mibi_phylo_object) %in% samples_to_remove),
  mibi_phylo_object
)
# Check the rarefaction curves:
#Requires: mibi_phylo_object
source("scripts/02_rarefaction_curves.R")
```
## 2.1:Data Pre-processing: Filtering out based on prevalence and abundance
```{r}

# --- Filtering parameters:USER INPUT SECTION ---
# Thresholds
prev_filter_threshold = 3   # Present in at least X samples
abund_filter_threshold = 5 # At least X reads 

# Logic Switches (1 = On, 0 = Off)
  # Here you can specify which filtering must be done,
  # Only based on prevalence, or abudance. Then fill in a 0 accordingly
  # Most of the time it is a combination of both prevalence and abundance, for example, at least >10 reads in 5 samples.
prev_filter = 1 
abund_filter = 1
# --------------------------
# Run the background script
# Requires: mibi_phylo_object, prev_filter_threshold, abund_filter_threshold...
source("scripts/02_filtering_prevalence_abundance.R")
#The script starts with taking the mibi_phylo_object and makes the: 
#The resulting object 'mibi_phylo_object_filt' is now ready 
```
## 2.1:Data Pre-processing: Filtering based on variance OPTIONAL, !!!This is not working yet!!!Also this should not be used for alpha or beta diversity, However it might be usefull before differential abundance analysis
```{r}
#Check the distribution of the varriances 
otu <- as.matrix(microbiome::abundances(mibi_phylo_object_filt))
taxon_variance <- apply(otu, 1, IQR)
taxon_variance <- apply(as.matrix(microbiome::abundances(mibi_phylo_object_filt)), 1, IQR)> quantile(taxon_variance, probs = seq(0, 1, 0.05))

#This will remove a lot of taxa, since there will be many taxa with 0 variance because they only occur in one sample.
# --- Filtering parameters: USER INPUT SECTION ---
# Variance Cutoff (e.g., 0.1 removes the bottom 10% least variable taxa)
variance_cutoff = 0.00

# Logic Switch (1 = On, 0 = Off)
variance_filter_switch = 1 
# --------------------------
# Run the background script
source("scripts/02_filtering_variance.R")
#This script starts with mibi_phylo_object_filt and updates it with this filtering 
# The resulting object 'mibi_phylo_object_filt' is now updated
```

## 2.1:Data Pre-processing: Sample removal or subset selection
```{r}
#Define the target group you want to subset
target_group = 'SC6'
# 1. Subset the samples
mibi_phylo_object_subset <- subset_samples(mibi_phylo_object_filt, SC_Number == target_group)

# 2. Corrected Pruning: Sum the counts within the NEW subset
# This ensures that if an ASV only existed in 'SC1', it is removed from the 'SC2' object.
mibi_phylo_object_subset <- prune_taxa(taxa_sums(mibi_phylo_object_subset) > 0, mibi_phylo_object_subset)

# 3. Reporting (Partners love to see how much was lost)
message(paste("Subset selected:", target_group))
message(paste("Samples remaining:", nsamples(mibi_phylo_object_subset)))
message(paste("Taxa remaining after pruning zeros:", ntaxa(mibi_phylo_object_subset)))
```
## 2.2 Object Finalization, selecting the object for the data analysis and normalisation
```{r}
## 2.3 Object Finalization
# Define which object flows into the analysis scripts
# Options: mibi_phylo_object_filt (full) or mibi_phylo_object_subset (subset)
ps_work <- mibi_phylo_object_filt
#str(ps_work)

```

2.2:Data Pre-processing: Normalisation
## 3.0 Normalization & The "Zero Problem"

Microbiome data is **compositional**. Because total read counts are an artifact of sequencer depth, they represent "parts of a whole" rather than absolute concentrations. If one taxon blooms, others appear to decrease simply because they occupy a smaller percentage of the total pool. 

### 3.1 Handling Sparsity (Zero Imputation)
Log-ratio transformations (like CLR) cannot handle zeros because $\log(0)$ is undefined. We address this using two distinct approaches before transformation:

* **Simple Pseudo-count (+1):** Adds a constant shift of 1 to every cell. It is fast and robust, though it can slightly bias ratios in extremely sparse datasets.
* **Bayesian-Multiplicative Replacement (BMR):** Uses the `zCompositions` package to impute zeros based on a Dirichlet prior. This is more mathematically rigorous as it preserves the ratio structure of the data.
    * **Note:** BMR requires careful handling of "singletons." We have adjusted the `z.delete` and `z.warning` thresholds to ensure samples with limited diversity are not dropped during the imputation process.
* **Matrix Completion:** *Planned for future implementation.*

---
### 3.2 Executed Transformations
The following background script processes the chosen `ps_work` object and generates the normalized data structures, using different techniques.
```{r}
# Requires: ps_work (defined in the data pre-processing section)
source("scripts/03_normalisation.R")

#The following objects are created:
#mibi_tss
#mibi_pseudo_clr
#mibi_bmr_clr
```
###The script generates three primary objects. Choosing the right one depends on your specific analysis:

-mibi_tss	Total Sum Scaling	Visualization: Relative Abundance (%) used for stacked bar plots.

-mibi_pseudo_clr	Pseudo-CLR (+1)	Statistics: Fast transformation for PCoA and basic linear modeling.

-mibi_bmr_clr	BMR-CLR (Bayesian)	Advanced Stats: High-fidelity modeling and differential abundance.

## 4.0 Alpha Diversity: Community Trait Analysis

Alpha diversity describes the complexity of a microbial community within a single sample. Rather than looking at "who is there," we look at the mathematical properties of the population structure. We categorize these traits into four pillars:

* **Richness:** The total number of unique taxa observed (how many species?).
* **Dominance:** Measures if the community is "top-heavy" with one or two abundant taxa.
* **Evenness:** How equally the individuals are distributed among the species.
* **Information:** Mathematical indices (like Shannon) that balance both richness and evenness.

> **Important:** Alpha diversity calculations are sensitive to normalization. We use the filtered, non-transformed object (`mibi_phylo_object_filt`) to ensure we are counting physical reads, not mathematical ratios.


```{r alpha_config}
# 1. Input Object (Filtered counts)
alfa_diversity_phyloseq_object <- mibi_phylo_object_filt

# 2. USER INPUT: Select Community Traits
selected_richness    <- c('observed', 'chao1') 
selected_dominance   <- c('absolute', 'dbp') 
selected_evenness    <- c('pielou')
selected_information <- c('shannon') 

# 4. Run Scripts
source("scripts/04_alpha_diversity.R")
# Preview results
knitr::kable(head(alpha_output), digits = 3)

#We visualize the distribution of community traits across our experimental groups. Note that the Y-axis scales are "free" because the units for Richness (counts) and Dominance (0-1) are fundamentally different.  

# 3. USER INPUT: Plotting Configuration
plot_x_axis   <- "SC_Number" #  What is on the X-axis? (e.g., "Timepoint", "group")
plot_color_by <- "SC_Number" # What defines the colors? (e.g., "group", "replicate")
plot_cols     <- 2              # Number of columns in the plot grid
plot_metrics <- plot_metrics_all
#plot_metrics <- c("observed", "shannon") # Use all calculated, or specific ones: c("observed", "shannon")
source("scripts/04_alpha_plots.R")

plot(alpha_boxplot)
```

## 5.0 Beta Diversity: Community Dissimilarity

Beta diversity quantifies the "distance" or "dissimilarity" between pairs of samples. To provide a non-redundant overview, we categorize these metrics based on whether they consider abundance (quantitative) vs. presence (qualitative), and phylogeny (evolutionary relatedness).

* **Aitchison:** Euclidean distance on CLR-transformed data. The gold standard for compositional data.
* **Bray-Curtis:** Non-phylogenetic; accounts for abundance.
* **Jaccard:** Non-phylogenetic; accounts only for presence/absence.
* **UniFrac (Weighted/Unweighted):** Phylogenetic; accounts for evolutionary distances between taxa.

> **Note:** For Aitchison, we use the CLR-imputed objects. For Bray, Jaccard, and UniFrac, we use Total Sum Scaling (TSS) to ensure sequencing depth doesn't bias the distances.

```{r, simple visualisation}
# --- USER INPUT: Beta Diversity Configuration ---
beta_metric = "wunifrac"    # Options: "aitchison", "bray", "jaccard", "unifrac", "wunifrac"
clr_variant = "BMR"         # If using Aitchison, select: "BMR" or "pseudo"
color_var   = "Groundwater_level"    # Metadata column for grouping/colors
shape_var   = NULL           # Optional: Metadata column for shapes (or NULL)
group_clustering <- "Groundwater_level"
shape_var <- "SC_Number"
# 1. Run Calculation (Ordination & PERMANOVA)
source("scripts/05_beta_diversity_calc.R")
# 2. Run Visualization
source("scripts/05_beta_diversity_visualisation.R")
# Display Plot and PERMANOVA results
plot(p_beta)
knitr::kable(permanova_res, caption = paste("PERMANOVA results for", beta_metric))

```
## 5.2 Advanced Ordination: Environmental Correlation PCA variant
#In this section, we pimp the ordination by mapping environmental variables (vectors) and group shapes (hulls) onto the microbial landscape.
```{r advanced_beta_vis PCA}
# --- USER INPUT: Beta Diversity Configuration ---
beta_metric      <- "aitchison" # only aitchison possible since it is PCA ordination
clr_variant      <- "BMR"
color_var        <- "Groundwater_level"
group_clustering <- "Groundwater_level"
shape_var        <- "SC_Number"

# 1. Run Calculation (Ordination & PERMANOVA)
source("scripts/05_beta_diversity_calc.R")
# PIMPING OPTIONS, for env fit:
head(ps_beta_input@sam_data)
env_variables    <- c("pH", "Av_Iron.J.kg.", "Conductivity", "SUM.PAHs..EPA.","total.benzene.oxidation.capacity..mg.kg.") # Continuous metadata
top_asv_n        <- 5    # Show top 5 ASVs driving the plot
target_level <- "order" #define taxonomic level which is given to the taxa, choose your level here: "genus", "family", "order", "class" "phylum", etc
# scalinglength of the arrows
arrow_reduction_taxa <- 5
arrow_reduction_env <- 2

# Run the advanced script
source("scripts/05_beta_advanced_vis_PCA.R")

# Display
plot(p_beta_PCA)
```
## 5.3 Advanced Ordination: Environmental Correlation PCoA variant
```{r advanced_beta_vis PCoA}
# --- USER INPUT: Beta Diversity Configuration ---
beta_metric      <- "wunifrac" # Or "bray", "jaccard", "unifrac" 
color_var        <- "Groundwater_level"
group_clustering <- "Groundwater_level"
shape_var        <- "SC_Number"
# 1. Run Calculation (Ordination & PERMANOVA)
source("scripts/05_beta_diversity_calc.R")

# PIMPING OPTIONS:
env_variables    <- c("pH", "Av_Iron.J.kg.", "Conductivity", "SUM.PAHs..EPA.","total.benzene.oxidation.capacity..mg.kg.")
top_asv_n        <- 5    # Show top 5 ASVs driving the plot
target_level <- "class" #define taxonomic level which is given to the taxa, choose your level here: "genus", "family", "order", "class" "phylum", etc
# scalinglength of the arrows
arrow_reduction_taxa <- 0.2
arrow_reduction_env <- 0.1

# Run the PCoA-specific visualization
source("scripts/05_beta_advanced_vis_PCoA.R")

# Display
plot(p_beta_PCoA)
```
## 6. Differential_abundance_LinDA.R
```{r, input script for LinDA}
## Select you object for analysis: it uses ps_work
# We use the filtered and not normalised phyloseq object, since LinDA handels the normalisation
# To make plots readable agglomerate at the selected phylogenetic level
# select the phylogenetic level to what the taxa should be agglomerated.
target_level_linda <- "family" # "species", "genus", "family", "order", "class" "phylum"
source("scripts/06_preparing_object_for_Linda.R")

# Set your primary comparison and reference group
main_var <- "Groundwater" # Adjust to match your exact metadata column name
ref_group <- "Below_GW" #ref_group <- "Below_GW" , depending on the levels of a specific factor.

# Set Reference
meta_df[[main_var]] <- relevel(factor(meta_df[[main_var]]), ref = ref_group)
# Define Formula: This model identifies which bacteria significantly change their abundance in response to the parameters defined in the model:
# Here: Main Effects + Interaction + Random Effect for replicates
# Grouping_SC = SC1, SC2, etc. (the sampling points)
linda_formula <- "~ Groundwater* SUM.PAHs..EPA. + (1 | SC_Number)"

#Run Linda:
# LinDA will drop these anyway, but it's good to be aware.
linda_res <- linda(
  feature.dat = otu_tab, 
  meta.dat = meta_df, 
  formula = linda_formula,
  feature.dat.type = "count",
  prev.filter = 0.10,        # Equivalent to your prev.cut
  is.winsor = TRUE,          # Enables outlier replacement
  outlier.pct = 0.03,        # Percentage of outliers to winsorize (default)
  zero.handling = "imputation", # imputation method, either 'pseudo-count' or imputation, exact imputation strategy can be found here: https://link.springer.com/article/10.1186/s13059-022-02655-5 
  pseudo.cnt = 0.5,
  alpha = 0.05,              # FDR significance level
  p.adj.method = "BH"        # Benjamini-Hochberg adjustment
)
#From this linda output you can plot multiple things
```

# Plotting the Linda results 
```{r}
## Plotting & Filtering
# Define which effect you want to visualize from names(linda_res$output)
# Example: "Groundwater_levelabove" or "SUM.PAHs..EPA." or the interaction
print(names(linda_res$output))
var_to_plot <- "GroundwaterAbove_GW" 
var_to_plot <- "SUM.PAHs..EPA." 
var_to_plot <- "GroundwaterAbove_GW:SUM.PAHs..EPA." 
# The interpretation of these results must be done carefully depending on the Linda input, Think about the model formula, and reference level since these will be essential in interpreting the output
# For example, if you are working with a model which includes an interaction term. If you specify the reference group as above groundwater, then if you plot the 'SUM.PAHs..EPA.' then you will get how the bacteria differ in log fold change as if the conditions were above groundwater. 
# Now if we are interested in how the bacteria differ in log fold changes with respect to PAH level below groundwater level, we need to change the reference group to below groundwater level to base the model on. 
# There is the possibility: "Groundwater_levelbelow:SUM.PAHs..EPA." which will indicate how the taxa 'respond to PAH levels below groundwater level' but the interaction term changes the interpretation greatly. So be carefull with the interpretation.
source("scripts/06_differential_abundance_LinDA.R")
```




