---
title: "Rmarkdown_16S_Ananlysis_mibibio"
author: "Simon Vandersanden"
date: "2026-03-05"
output:
  html_document: default
  pdf_document: default
params:
  config_file: "configs/Config_Griftpark_Wetland_All_Wetland_Samples_New_Metadata.yaml"
---

```{r setup, include=FALSE}
#working with a configuration file, to make the workflow more flexible.
library(yaml)
# SAFETY NET: If params doesn't exist (manual run), manually point to the file
if (!exists("params") || is.null(params$config_file)) {
  config_path <- "configs/Config_Griftpark_Wetland_All_Wetland_Samples_New_Metadata.yaml"
} else {
  config_path <- params$config_file
}
config <- yaml::read_yaml(config_path)
# This is the lasted up to date config file:Config_Griftpark_Wetland_All_Wetland_Samples_New_Metadata.yaml
```


## Loading in the necessary libraries for basic handeling

```{r, results= 'hide'}
library(phyloseq)
library(dplyr)
library(picante)    # For Phylogenetic Diversity
library(microbiome) # For Dominance metrics
library(vegan)      # For Rarefaction curves
library(zCompositions)
library(ggplot2)
library(ggrepel)
library(MicrobiomeStat)
library(scales)
library(paletteer)
library(stringr)
library(yaml)
library(NetCoMi)
set.seed(99)
```

## 1:Loading in and inspecting the different parts of the phyloseq object
Here you load in you microbiome data, there is a part where sample data can be linked to the object, so you can quickly add in new sample data for analysis and plotting.


```{r,include=TRUE, results='hide'}
# --- 1. LOAD PHYLOSEQ OBJECT ---
# This uses the path defined in the YAML file.
mibi_phylo_object <- readRDS(config$paths$phyloseq_rds) 
# Extract tables for inspection
mibi_otu_table  <- as(otu_table(mibi_phylo_object), "matrix")
mibi_tax_table  <- as(tax_table(mibi_phylo_object), "matrix")
mibi_meta_table <- as(sample_data(mibi_phylo_object), "data.frame")

# --- 2. CLEAN TAXONOMY ---
# Instead of "Unclassified", we use the variable from your config
uncl_label <- config$analysis_params$unclassified_label

mibi_tax_table[is.na(mibi_tax_table)] <- uncl_label
mibi_tax_table[mibi_tax_table == ""]  <- uncl_label

# Update the phyloseq object with cleaned taxonomy
tax_table(mibi_phylo_object) <- tax_table(mibi_tax_table)

# --- 3. UPDATE METADATA ---
# We replace the filename, the separator (\t), and the decimal (,) with config values
mibi_meta_table_new <- read.table(
  file = config$paths$metadata_txt, 
  header = TRUE, 
  sep = config$analysis_params$metadata_sep, 
  row.names = 1, 
  stringsAsFactors = FALSE, 
  dec = config$analysis_params$metadata_dec
)

# Safety check
stopifnot(all(sample_names(mibi_phylo_object) %in% rownames(mibi_meta_table_new)))

# Link the new metadata to the object
sample_data(mibi_phylo_object) <- mibi_meta_table_new
```

## 2.1:Data Pre-processing: Filtering out all non-samples and checking rarefaction curves
Normally some pre-filtering has been done on the reads before they were put into the phyloseq object: the sequencing depth threshhold, the decontamination, and some taxonomic filtering of certain undesired taxa such as Eukaryota, Mitochondria and Chloroplast sequences.
### Here we can remove some unecessary samples from the dataset.
### There is also the possibility to plot the rarefaction curves, to see wheter the sequencing depth is sufficiënt.

```{r,include=TRUE}
# It is possible that that some blancs and mock samples are still present in the samples, these should be filtered out because we will only filter using all the 'real samples'
# Filter out unwanted samples using the list from the config
samples_to_remove <- config$filtering$samples_to_remove

mibi_phylo_object <- prune_samples(
  !(sample_names(mibi_phylo_object) %in% samples_to_remove),
  mibi_phylo_object
)
# Run the rarefaction script using the path from config
#source(config$scripts$rarefaction)
```
## 2.1:Data Pre-processing: Filtering out based on prevalence and abundance

Often prevalence or abundance filtering is done to remove uncertain features which can be artifacts created. or remove rare sequences, which might not be of interest to us. This is highly customisable so also all features can be kept however often for clarity, features are filtered out.
Here you can filter on prevalence and abundance, you can also select either prevalence or either abundance.

```{r}
config <- yaml::read_yaml(config_path)
# --- Filtering parameters: PULLED FROM CONFIG ---
prev_filter_threshold  <- config$filtering$prev_filter_threshold
abund_filter_threshold <- config$filtering$abund_filter_threshold

# Logic Switches
# We convert the YAML boolean (true/false) to your script's expected 1/0
prev_filter  <- as.numeric(config$filtering$apply_prevalence)
abund_filter <- as.numeric(config$filtering$apply_abundance)

# Run the background script
# Note: Ensure these scripts use the variables defined above!
source(config$scripts$filtering)

# Resulting object 'mibi_phylo_object_filt' is now ready
```

## 2.1:Data Pre-processing: Sample removal or subset selection
If you have a large dataset or are interested in specific samples, here you can select which samples or groups of samples you want to take a closer look at.
### The script knows whether to subset or keep the full dataset based on the config file
```{r}
config <- yaml::read_yaml(config_path)
if (config$subsetting$use_subset) {
  
  target_group <- config$subsetting$target_value
  target_col   <- config$subsetting$column_name
  
  # Professional trick: use 'get' to find the column name dynamically from the string in YAML
  mibi_phylo_object_subset <- subset_samples(mibi_phylo_object_filt, 
                                             get(target_col) == target_group)
  
  # Remove taxa with zero counts in this specific subset
  mibi_phylo_object_subset <- prune_taxa(taxa_sums(mibi_phylo_object_subset) > 0, 
                                         mibi_phylo_object_subset)
  
  # Reporting
  message(paste("Subset selected:", target_group))
  message(paste("Samples remaining:", nsamples(mibi_phylo_object_subset)))
  message(paste("Taxa remaining after pruning zeros:", ntaxa(mibi_phylo_object_subset)))
  
} else {
  message("No subsetting applied. Using full filtered dataset.")
  mibi_phylo_object_subset <- mibi_phylo_object_filt
}
```
## 2.2 Object selection: selecting the object for the normalisation, and further data analysis.
```{r}
# Define which object flows into the analysis scripts.
# We use the 'subset' object, which is either the actual subset 
# OR the full dataset (if no subsetting was applied in the previous chunk).
ps_work <- mibi_phylo_object_subset
#Change the sample names to logical names
sample_names(ps_work) <- ps_work@sam_data$sample_names
# --- Reporting: Final Object Overview ---
# 1. Get the total counts
n_samples <- nsamples(ps_work)
n_taxa    <- ntaxa(ps_work)
# 2. Get the names and collapse them into one string
sample_list_string <- paste(sample_names(ps_work), collapse = ", ")

# 3. Summary of the samples used for downstream analysis
message("--------------------------------------------------")
message(paste("PROJECT:", config$project_metadata$name))
message(paste("ANALYSIS TARGET:", if(config$subsetting$use_subset) config$subsetting$target_value else "Full Dataset"))
message(paste("TOTAL SAMPLES:", n_samples))
message(paste("TOTAL TAXA:", n_taxa))
message("SAMPLES INCLUDED:")
message(sample_list_string)
message("--------------------------------------------------")

```

# 2.2:Data Pre-processing: Normalisation & The "Zero Problem"

Microbiome data is **compositional**. Because total read counts are an artifact of sequencer depth, they represent "parts of a whole" rather than absolute concentrations. If one taxon blooms, others appear to decrease simply because they occupy a smaller percentage of the total pool.

### 3.1 Handling Sparsity (Zero Imputation)

Log-ratio transformations (like CLR) cannot handle zeros because $\log(0)$ is undefined. We address this using two distinct approaches before transformation:

-   **Simple Pseudo-count (+1):** Adds a constant shift of 1 to every cell. It is fast and robust, though it can slightly bias ratios in extremely sparse datasets.
-   **Bayesian-Multiplicative Replacement (BMR):** Uses the `zCompositions` package to impute zeros based on a Dirichlet prior. This is more mathematically rigorous as it preserves the ratio structure of the data.
    -   **Note:** BMR requires careful handling of "singletons." We have adjusted the `z.delete` and `z.warning` thresholds to ensure samples with limited diversity are not dropped during the imputation process.
-   **Matrix Completion:** *Planned for future implementation.*

------------------------------------------------------------------------
### 3.2 Executed Transformations
The following background script processes the chosen `ps_work` object and generates the normalized data structures, using different methods.
```{r}
# Requires: ps_work (defined above)
# This script will generate: mibi_tss, mibi_pseudo_clr, mibi_bmr_clr

source(config$scripts$normalization)

# Quick check to ensure the objects were created
if(exists("mibi_tss")) {
  message("Normalization complete: mibi_tss, mibi_pseudo_clr, and mibi_bmr_clr are ready.")
}
#quickly save the generated objects for a quicker later analysis
saveRDS(mibi_tss, file = "Objects_Generated_In_Pipeline/mibi_tss_All_Wetland_Samples.rds")
saveRDS(mibi_pseudo_clr, file = "Objects_Generated_In_Pipeline/mibi_pseudo_clr_All_Wetland_Samples.rds")
saveRDS(mibi_bmr_clr, file = "Objects_Generated_In_Pipeline/mibi_bmr_clr_All_Wetland_Samples.rds")
saveRDS(ps_work, file = "Objects_Generated_In_Pipeline/ps_work_All_Wetland_Samples.rds")
```

###The script generates three primary objects. Choosing the right one depends on your specific analysis:

-mibi_tss Total Sum Scaling Visualization: Relative Abundance (%) used for stacked bar plots.

-mibi_pseudo_clr Pseudo-CLR (+1) Statistics: Fast transformation for PCoA and basic linear modeling.

-mibi_bmr_clr BMR-CLR (Bayesian) Advanced Stats: High-fidelity modeling and differential abundance.

Specific for the wetland sample analysis, a chunk of code to make some adjustments to the objects
```{r}
# We replace the filename, the separator (\t), and the decimal (,) with config values
mibi_meta_table_new <- read.table(
  file = config$paths$metadata_txt, 
  header = TRUE, 
  sep = config$analysis_params$metadata_sep, 
  row.names = 1, 
  stringsAsFactors = FALSE, 
  dec = config$analysis_params$metadata_dec
)

# Load in the previously saved and normalized objects
mibi_tss <- readRDS("Objects_Generated_In_Pipeline/mibi_tss_All_Wetland_Samples.rds") 
mibi_pseudo_clr <-readRDS("Objects_Generated_In_Pipeline/mibi_pseudo_clr_All_Wetland_Samples.rds") 
mibi_bmr_clr <- readRDS("Objects_Generated_In_Pipeline/mibi_bmr_clr_All_Wetland_Samples.rds") 
ps_work <- readRDS("Objects_Generated_In_Pipeline/ps_work_All_Wetland_Samples.rds") 
# Load in the previously saved and normalized objects all samples included
#mibi_tss <- readRDS("Objects_Generated_In_Pipeline/mibi_tss_All_Samples.rds") 
#mibi_pseudo_clr <-readRDS("Objects_Generated_In_Pipeline/mibi_pseudo_clr_All_Samples.rds") 
#mibi_bmr_clr <- readRDS("Objects_Generated_In_Pipeline/mibi_bmr_clr_All_Samples.rds") 
#ps_work <- readRDS("Objects_Generated_In_Pipeline/ps_work_All_Samples.rds") 

# Change the samples names, so they match with the metadata formatting
sample_names(ps_work) <- ps_work@sam_data$Sample_ID
sample_names(mibi_tss) <- mibi_tss@sam_data$Sample_ID
sample_names(mibi_pseudo_clr) <- mibi_pseudo_clr@sam_data$Sample_ID
sample_names(mibi_bmr_clr) <- mibi_bmr_clr@sam_data$Sample_ID
#load in the updated metadata table 
sample_data(ps_work) <- mibi_meta_table_new
sample_data(mibi_tss) <- mibi_meta_table_new
sample_data(mibi_pseudo_clr) <- mibi_meta_table_new
sample_data(mibi_bmr_clr) <- mibi_meta_table_new

# Rename them again for interpretation:  Change the samples names, so they are based on the sample names:
sample_names(ps_work) <- ps_work@sam_data$sample_names
sample_names(mibi_tss) <- mibi_tss@sam_data$sample_names
sample_names(mibi_pseudo_clr) <- mibi_pseudo_clr@sam_data$sample_names
sample_names(mibi_bmr_clr) <- mibi_bmr_clr@sam_data$sample_names


### Make subsets in ps_work per Constructed wetland for specific analysis:
barchart_target   <- "Wetland" #
barchart_target_col <- "wetland_sample" #
# Professional trick: use 'get' to find the column name dynamically from the string in YAML
ps_work <- subset_samples(ps_work, get(barchart_target_col) == barchart_target)
mibi_bmr_clr <- subset_samples(mibi_bmr_clr, get(barchart_target_col) == barchart_target)
mibi_tss <- subset_samples(mibi_tss, get(barchart_target_col) == barchart_target)


```

# Community visualisation: barplots
This chunk is designed to plot relative abundances of you microbial community data, this takes the microbial community and you can select at which level the plotting is done, the labeling is two levels to ensure comprehensiveness, and detail (higher phylogenetic level and lower for the detail). The main color is determined by the high level taxonomy, variations of this main color are used in the lower levels. Several choices can be made, a relative abundance threshold level, which is a cutoff level for relative abundance, if a taxon does not meet this threshold i.e. all taxa with relative abundance below this threshold are put into 'Others'
```{r plot abundance, fig.width=12, fig.height=8}
config <- yaml::read_yaml(config_path)
# --- SETUP ---
source(config$scripts$barplot_main)
# --- 1. INPUT FROM CONFIG ---
v_glob     <- config$Barplot_specific$global_barplots
pal_choice <- config$Barplot_specific$palette_choice

# --- 2. PREPARATION ---
data_for_plot <- prepare_abundance_data(
  ps         = ps_work,
  high_level = v_glob$high_level,
  low_level  = v_glob$low_level, 
  facet_var  = v_glob$facet_var, 
  threshold  = v_glob$threshold 
)
# This prepared the colours, so custom colours are assigned to the created groups depending on the taxonomic specification.
#Run this once
#set

#my_colors <- get_shaded_palette(data_for_plot, pal_choice)

# --- 3. DYNAMIC PLOTTING ---
# This is a complex plotting command that is constructed to take into acount the previously specified parameters, high and low taxonomic levels, relative abundance threshold cutoff, how to group the samples for plotting. So this information is also found on the finally generated plot

Relative_Abundance_Plot <- ggplot(data_for_plot, aes(x = !!sym(v_glob$facet_var), y = Abundance, fill = Hierarchical_Label)) +
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.05) +
  scale_fill_manual(values = my_colors) +
  labs(
    title = paste("Relative Abundance:", str_to_title(v_glob$high_level), "-", str_to_title(v_glob$low_level)),
    subtitle = paste("Filtered at", v_glob$threshold, "% threshold | Grouped by", v_glob$facet_var),
    x = stringr::str_replace_all(v_glob$facet_var, "_", " "),
    y = "Relative Abundance (%)",
    fill = paste(str_to_title(v_glob$high_level), "-", str_to_title(v_glob$low_level))
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.text = element_text(size = 9),
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold")
  ) +
  guides(fill = guide_legend(ncol = 2))

Relative_Abundance_Plot

# Save output using config values for naming
ggsave(paste0("results/CW_3/Relative_Abundance_Plot_", v_glob$high_level, "_", v_glob$low_level, ".png"), 
       plot = Relative_Abundance_Plot, width = v_glob$fig_width, height = v_glob$fig_height, dpi = v_glob$dpi)
```

## Community visualisation, detailed group analysis.
This chunk of code is designed to plot relative abundances of you microbial community data. But here you can look within specific group and get the lower level taxonomic distribution so you can gain insights into specific important groups within you data.The labeling is two levels to ensure comprehensiveness, and detail (higher phylogenetic level and lower for the detail). The main color is determined by the high level taxonomy, variations of this main color are used in the lower levels. Several choices can be made, a relative abundance threshold level, which is a cutoff level for relative abundance, if a taxon does not meet this threshold i.e. all taxa with relative abundance belower this threshold are put into 'Others' 
```{r plot specific group abundance, fig.width=12, fig.height=8}
source(config$scripts$barplot_focus)
# --- 1. DEFINE SPECIFICS FROM CONFIG ---
v_foc      <- config$Barplot_specific$focus_barplot
pal_choice <- config$Barplot_specific$palette_choice

# From this specifically selected group there is the possibility to plot the lower taxonomic levels within it, similar logic to a normal relative abundance plot, but now the relative abundance within this selected taxonomic group is plotted.

# --- 2. PREPARATION ---
data_subset <- prepare_specific_abundance_data(
  ps         = ps_work,
  rank_name  = v_foc$target_rank,
  rank_value = v_foc$target_val,
  high_level = v_foc$high_level,
  low_level  = v_foc$low_level,
  facet_var  = v_foc$facet_var,
  threshold  = v_foc$threshold
)

# This prepared the colors, so custom colors are assigned to the created groups depending on the taxonomic specification.
my_colors_sub <- get_shaded_palette(data_subset, pal_choice)# Uses the shared function from: 08_barchart_visualisation.R

# --- 3. DYNAMIC PLOTTING ---
# This is a complex plotting command that is constructed to take into acount the previously specified parameters, high and low taxonomic levels, relative abundance threshold cutoff, how to group the samples for plotting. So this information is also found on the finally generated plot
Specific_Group_Relative_Abundance_Plot <- ggplot(data_subset, aes(x = !!sym(v_foc$facet_var), y = Abundance, fill = Hierarchical_Label)) +
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.05) +
  scale_fill_manual(values = my_colors_sub) +
  labs(
    title = paste("Focus:", v_foc$target_val, "(", str_to_title(v_foc$high_level), "-", str_to_title(v_foc$low_level), ")"),
    subtitle = paste("Filtered at", v_foc$threshold, "% internal abundance | Grouped by", v_foc$facet_var),
    x = stringr::str_replace_all(v_foc$facet_var, "_", " "),
    y = paste("Relative Abundance within", v_foc$target_val, "(%)"),
    fill = paste(str_to_title(v_foc$high_level), "-", str_to_title(v_foc$low_level))
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.text = element_text(size = 8),
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold", color = "darkblue")
  ) +
  guides(fill = guide_legend(ncol = 2))

Specific_Group_Relative_Abundance_Plot

ggsave(paste0("results/Specific_Group_Relative_Abundance_Plot_", v_foc$target_val, ".png"), 
       plot = Specific_Group_Relative_Abundance_Plot, width = v_foc$fig_width, height = v_foc$fig_height, dpi = v_foc$dpi)
```

## 4.0 Alpha Diversity: Community Trait Analysis

Alpha diversity describes the complexity of a microbial community within a single sample. Rather than looking at "who is there," we look at the mathematical properties of the population structure. We categorize these traits into four pillars:

-   **Richness:** The total number of unique taxa observed (how many species?).
-   **Dominance:** Measures if the community is "top-heavy" with one or two abundant taxa.
-   **Evenness:** How equally the individuals are distributed among the species.
-   **Information:** Mathematical indices (like Shannon) that balance both richness and evenness.

> **Important:** Alpha diversity calculations are sensitive to normalization. We use the filtered, non-transformed object (`mibi_phylo_object_filt`) to ensure we are counting physical reads, not mathematical ratios.

```{r plot different alpha diversity measures, fig.width=12, fig.height=8}
config <- yaml::read_yaml(config_path)
# --- 1. SETUP & INPUT ---
# Use the ps work object
alfa_diversity_phyloseq_object <- ps_work

# Pull traits from Config
selected_richness    <- config$Alpha_Diversity$traits$richness
selected_dominance   <- config$Alpha_Diversity$traits$dominance
selected_evenness    <- config$Alpha_Diversity$traits$evenness
selected_information <- config$Alpha_Diversity$traits$information

# --- 2. RUN CALCULATION SCRIPT ---
# This script uses the variables above to generate 'plot_metrics_all'
source(config$scripts$alpha_calc)

# --- 3. PLOTTING CONFIGURATION ---
plot_x_axis   <- config$Alpha_Diversity$plotting$x_axis
plot_color_by <- config$Alpha_Diversity$plotting$color_by
plot_cols     <- config$Alpha_Diversity$plotting$grid_cols



# Logic to handle 'all' vs specific metrics
if (config$Alpha_Diversity$plotting$selected_metrics == "all") {
  plot_metrics <- plot_metrics_all
} else {
  plot_metrics <- config$Alpha_Diversity$plotting$selected_metrics
}


# Define your strict logical order here
#desired_order <- c("Section_1", "Section_2", "Section_3", "Section_4", "Section_5") # Change this based on the variables used for

# --- CUSTOMISATION BEFORE ALPHA PLOTTING ---
# Automatically apply the ordering to the X-axis column if found in alpha_output
#if (plot_x_axis %in% colnames(alpha_output)) {
#  alpha_output[[plot_x_axis]] <- factor(alpha_output[[plot_x_axis]], levels = desired_order)
#}

# Automatically apply the ordering to the Color column if found in alpha_output
#if (plot_color_by %in% colnames(alpha_output)) {
#  alpha_output[[plot_color_by]] <- factor(alpha_output[[plot_color_by]], levels = desired_order)
# --- 4. RUN PLOTTING SCRIPT ---

source(config$scripts$alpha_plot)

# Display and Save
plot(alpha_boxplot)
#We visualize the distribution of community traits across our experimental groups. Note that the Y-axis scales are "free" because the units for Richness (counts) and Dominance (0-1) are fundamentally different.  
ggsave(
  filename = "results/alpha_diversity_plot.png", 
  plot = alpha_boxplot, 
  width = config$Alpha_Diversity$plotting$fig_width, 
  height = config$Alpha_Diversity$plotting$fig_height, 
  dpi = config$Alpha_Diversity$plotting$dpi
)
```

## 5.0 Beta Diversity: Community Dissimilarity

Beta diversity quantifies the "distance" or "dissimilarity" between pairs of samples. To provide a non-redundant overview, we categorize these metrics based on whether they consider abundance (quantitative) vs. presence (qualitative), and phylogeny (evolutionary relatedness).

-   **Aitchison:** Euclidean distance on CLR-transformed data. The gold standard for compositional data.
-   **Bray-Curtis:** Non-phylogenetic; accounts for abundance.
-   **Jaccard:** Non-phylogenetic; accounts only for presence/absence.
-   **UniFrac (Weighted/Unweighted):** Phylogenetic; accounts for evolutionary distances between taxa.

> **Note:** For Aitchison, we use the CLR-imputed objects. For Bray, Jaccard, and UniFrac, we use Total Sum Scaling (TSS) to ensure sequencing depth doesn't bias the distances.

```{r plot different beta diversity measures, fig.width=12, fig.height=8}
config <- yaml::read_yaml(config_path)
# --- 1. CONFIG INPUT ---
beta_metric      <- config$Beta_Diversity$basic$metric
clr_variant      <- config$Beta_Diversity$basic$clr_variant
color_var        <- config$Beta_Diversity$color_var
shape_var        <- config$Beta_Diversity$shape_var
group_clustering <- config$Beta_Diversity$group_clustering
show_ellipses    <- isTRUE(config$Beta_Diversity$show_ellipses)
show_labels      <- isTRUE(config$Beta_Diversity$show_labels)

# --- 2. EXECUTION ---
source(config$scripts$beta_calc)

source(config$scripts$beta_vis_basic)

plot(p_beta)
```
## 5.2 Advanced Ordination: Environmental Correlation PCA variant

#In this section, we upgrade the ordination by mapping environmental variables (vectors) and group shapes (hulls) onto the microbial landscape.

```{r plot beta diversity measures PCA using aitchison distance, fig.width=12, fig.height=8}
config <- yaml::read_yaml(config_path)
# --- USER INPUT: Beta Diversity Configuration ---
# --- 1. CONFIG INPUT ---
beta_metric      <- config$Beta_Diversity$basic$metric
clr_variant      <- config$Beta_Diversity$basic$clr_variant
color_var        <- config$Beta_Diversity$color_var
shape_var        <- config$Beta_Diversity$shape_var
group_clustering <- config$Beta_Diversity$group_clustering
show_ellipses    <- isTRUE(config$Beta_Diversity$show_ellipses)
show_labels      <- isTRUE(config$Beta_Diversity$show_labels)
labels           <- config$Beta_Diversity$labels

# Advanced parameters (FIXED parsing paths)

numeric_env_variables            <- config$Beta_Diversity$advanced$numeric_env_variables
categ_env_variables         <- config$Beta_Diversity$advanced$categ_env_variables
top_asv_n                      <- config$Beta_Diversity$advanced$top_asv_n
target_level                   <- config$Beta_Diversity$advanced$target_tax_level
filter_taxon_loadings          <- isTRUE(config$Beta_Diversity$advanced$filter_taxon_loadings)
filter_taxon_abundance_filter  <- config$Beta_Diversity$advanced$filter_taxon_abundance_filter
filter_taxon_prevalence_filter <- config$Beta_Diversity$advanced$filter_taxon_prevalence_filter
arrow_reduction_taxa           <- config$Beta_Diversity$advanced$pca_arrow_taxa
arrow_reduction_env            <- config$Beta_Diversity$advanced$pca_arrow_env
arrow_env_permanova_significance <- 0.05

# --- 2. EXECUTION ---
source(config$scripts$beta_calc)
source(config$scripts$beta_vis_pca)

plot(p_beta_PCA)
```

## 5.3 Advanced Ordination: Environmental Correlation PCoA variant

```{r plot beta diversity measures PCoA using various distance matrices as input, fig.width=12, fig.height=8}
config <- yaml::read_yaml(config_path)
# --- USER INPUT: Beta Diversity Configuration ---
# --- 1. CONFIG INPUT ---
beta_metric      <- config$Beta_Diversity$basic$metric # Usually wunifrac/bray
color_var        <- config$Beta_Diversity$color_var
shape_var        <- config$Beta_Diversity$shape_var
group_clustering <- config$Beta_Diversity$group_clustering
show_ellipses    <- isTRUE(config$Beta_Diversity$show_ellipses)
show_labels      <- isTRUE(config$Beta_Diversity$show_labels)
labels           <- config$Beta_Diversity$labels
# Advanced parameters (PCoA specific scaling)
numerical_env_vars            <- config$Beta_Diversity$advanced$numerical_env_vars
numeric_env_variables            <- config$Beta_Diversity$advanced$numeric_env_variables
categorical_env_vars          <- config$Beta_Diversity$advanced$categorical_env_vars

numeric_env_variables            <- config$Beta_Diversity$advanced$numeric_env_variables
categ_env_variables         <- config$Beta_Diversity$advanced$categ_env_variables
top_asv_n            <- config$Beta_Diversity$advanced$top_asv_n
target_level             <- config$Beta_Diversity$advanced$target_tax_level
arrow_reduction_taxa <- config$Beta_Diversity$advanced$pcoa_arrow_taxa
arrow_reduction_env  <- config$Beta_Diversity$advanced$pcoa_arrow_env
fig_width <- config$Beta_Diversity$fig_width
fig_height <- config$Beta_Diversity$fig_height

# --- 2. EXECUTION ---
source(config$scripts$beta_calc)

#Customisation before plotting :
#desired_order <- c( "INF","EFF", "MF1", "MF2", "MF5", "MF6", "MF9", "MF10") # Change this based on the variables used for plotting
# Check if shape_var column exists in the phyloseq metadata and refactor it
#if (shape_var %in% colnames(phyloseq::sample_data(ps_beta_input))) {
#  # 1. Pull the metadata out
#  meta_tmp <- data.frame(phyloseq::sample_data(ps_beta_input))
#  # 2. Convert your target column to the ordered factor
#  meta_tmp[[shape_var]] <- factor(meta_tmp[[shape_var]], levels = desired_order)
#  # 3. Put it right back into the active phyloseq object
#  phyloseq::sample_data(ps_beta_input) <- phyloseq::sample_data(meta_tmp)
#}

source(config$scripts$beta_vis_pcoa)

plot(p_beta_PCoA)
```


#RDA
```{r}
ggplot() +
  # A. Origin Crosshairs
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray75", linewidth = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray75", linewidth = 0.5) +
  
  # B. Microbial Samples (Points) - Colored by Wetland, shaped by Depth
  geom_point(data = sample_coords, 
             aes(x = RDA1, y = RDA2, color = CW_Number, shape = Sample_depth_information), 
             size = 2.8, alpha = 0.75) +
  
  # C. Continuous Chemistry (Arrows)
  geom_segment(data = arrow_coords,
               aes(x = 0, y = 0, xend = RDA1 * arrow_scale, yend = RDA2 * arrow_scale),
               arrow = arrow(length = unit(0.20, "cm")), 
               color = "firebrick3", linewidth = 0.7, alpha = 0.8) +
  
  # D. Labels for Chemistry Arrows
  geom_text(data = arrow_coords,
            aes(x = RDA1 * (arrow_scale + 0.2), y = RDA2 * (arrow_scale + 0.2), label = Variable),
            color = "firebrick4", fontface = "bold", size = 3.5) +
  
  # E. Categorical Feature Centroids (Labels in white boxes)
  geom_label(data = centroid_coords,
             aes(x = RDA1 * arrow_scale, y = RDA2 * arrow_scale, label = Factor_Level),
             color = "dodgerblue4", fill = "white", fontface = "bold", size = 3.5,
             label.padding = unit(0.15, "lines"), box.padding = 0.5, alpha = 0.9) +
  
  # F. Theme & Aesthetics
  theme_bw() +
  scale_color_brewer(palette = "Set1") + 
  labs(title = "Robust Aitchison RDA Ordination Biplot",
       subtitle = "Model Adj. R² = 12.4%, Global p = 0.001 (All marginal terms p < 0.05)",
       x = "RDA1 (6.78% variance explained)", 
       y = "RDA2 (4.21% variance explained)",
       color = "Constructed Wetland",
       shape = "Sample Depth (m)") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        axis.title = element_text(face = "bold", size = 11),
        legend.title = element_text(face = "bold"),
        legend.position = "right")




```


## 6. Differential_abundance_LinDA.R
### We use the filtered and not normalised phyloseq object, since LinDA handels the normalisation
### To make plots readable agglomerate at the selected phylogenetic level
### select the phylogenetic level to what the taxa should be agglomerated.
In this script we do a differential abundance analysis. In the following chunk calculations are made.
```{r, input script for LinDA here the model is defined and differental abundance calculations are done}
config <- yaml::read_yaml(config_path)
# --- 1. PREPARATION ---
target_level_linda <- config$Differential_Abundance$target_level

# In order te get a readable plot for LinDA, it is better to agglomerate the reads to a specific phylogenetic level
# they you get information on how mayor groups of bacteria respond to the specified model, 
# otherwise raw sequences will be given.
#-----------------------------------------------------------------------------------------
ps_glommed <- phyloseq::tax_glom(ps_work, taxrank = target_level_linda, NArm = TRUE) 
#----------------------------------------------------------------------------------------
#In this step all ASVs which have the same taxonomic level will be agglomerated.
# For plotting we need to change the sequences in the OTU table to a specific taxonomic label otherwise raw sequences will be visible on the plot.
otu_tab <- as.data.frame(t(otu_table(ps_glommed)))
# extract the OTU table from the agglomerated object
tax_tab <- as.data.frame(tax_table(ps_glommed)) # extract the taxonomy table from the agglomerated object
taxa_labels <- as.character(tax_tab[, target_level_linda])
unique_taxa_ids <- make.unique(taxa_labels, sep = ".")
#is then added to the otu and tax table in the following line: 
rownames(otu_tab) <- unique_taxa_ids
rownames(tax_tab) <- unique_taxa_ids

#-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
#METADATA FORMATTING
#extract the metadata
metadata_correct_format <- data.frame(sample_data(ps_glommed))
#convert it to a suitable format
# Define your variables of interest
categ_env_variables   <- config$Differential_Abundance$categ_env_variables
numeric_env_variables <- config$Differential_Abundance$numeric_env_variables
# =====================================================================
# 1. AUTOMATED FACTOR CONVERSION LAYER (Generalized)
# =====================================================================
for (cat_var in categ_env_variables) {
  if (cat_var %in% colnames(metadata_correct_format)) {
    # Clean up column to a character format first, then cast to a standard factor
    # This ensures R handles any messy mixed-type inputs or text labels cleanly
    metadata_correct_format[[cat_var]] <- factor(as.character(metadata_correct_format[[cat_var]]))
    message(paste("Successfully converted to factor:", cat_var))
  } else {
    warning(paste("Configured categorical variable not found in dataset columns:", cat_var))
  }
}
# =====================================================================
# 2. AUTOMATED NUMERICAL CONVERSION LAYER (Generalized)
# =====================================================================
for (num_var in numeric_env_variables) {
  if (num_var %in% colnames(metadata_correct_format)) {
    # Step A: Convert to character and strip common hidden spacing/text issues
    clean_chars <- as.character(metadata_correct_format[[num_var]])
    clean_chars <- trimws(clean_chars) # Remove leading/trailing whitespaces
    # Step B: Cast explicitly to numeric double
    # Suppress warnings temporarily so R doesn't flood the console if 
    # text like "missing" or "BDL" naturally turns into NA
    suppressWarnings({
      numeric_vector <- as.numeric(clean_chars)
    })
    # Save the clean numeric vector back into your corrected formatting table
    metadata_correct_format[[num_var]] <- numeric_vector
    message(paste("Successfully converted to numeric metric:", num_var))
    # Optional Safety Check: Report if the conversion introduced any new NAs
    raw_na_count <- sum(is.na(clean_chars)) # check NAs before parsing
    new_na_count <- sum(is.na(numeric_vector))
    if (new_na_count > raw_na_count) {
      warning(paste0(
        "Variable '", num_var, "' contained non-numeric text strings (e.g., 'ND', '<LOD', or blanks) ",
        "which have been automatically forced to NA values."
      ))
    }
  } else {
    warning(paste("Configured numeric variable not found in dataset columns:", num_var))
  }
}
all_target_cols <- c(categ_env_variables, numeric_env_variables)
#-----------------------------------------------------------------------------
# Extract ONLY target metadata columns into a fresh table
linda_metadata <- metadata_correct_format[, all_target_cols]
# Optional check the structure of the final metadata object 
str(linda_metadata)


#------------------------------------------------------------------------------------------------------------------
# Setting up the model
config <- yaml::read_yaml(config_path)
# --- 2. MODEL SETUP ---
l_mod <- config$Differential_Abundance$model

# Ensure the formula is a single string (R sometimes imports YAML as a list)
linda_formula_string <- paste(l_mod$formula, collapse = "")

# Set Reference dynamically
linda_metadata [[l_mod$main_var]] <- relevel(factor(linda_metadata [[l_mod$main_var]]), 
                                     ref = l_mod$ref_group)

# 3. Ensure your random intercept is strictly classified as a factor
linda_metadata$CW_number_MF_number <- factor(linda_metadata$CW_number_MF_number)

# --- 3. RUN LINDA ---
l_params <- config$Differential_Abundance$params
linda_res <- linda(
  feature.dat     = otu_tab, 
  meta.dat        = linda_metadata, 
  formula         = linda_formula_string,
  feature.dat.type = "count",
  prev.filter     = l_params$prev_filter,
  is.winsor       = TRUE,
  outlier.pct     = l_params$outlier_pct,
  zero.handling   = l_params$zero_handling,
  pseudo.cnt      = l_params$pseudo_cnt,
  alpha           = l_params$alpha,
  p.adj.method    = l_params$p_adj_method
)
# From this linda output you can plot multiple things

```

# Plotting the Linda results
The interpretation of these results must be done carefully depending on the Linda input, Think about the model formula, and reference level since these will be essential in interpreting the output

For example, if you are working with a model which includes an interaction term. If you specify the reference group as above groundwater, then if you plot the 'SUM.PAHs..EPA.' then you will get how the bacteria differ in log fold change as if the conditions were above groundwater. 

Now if we are interested in how the bacteria differ in log fold changes with respect to PAH level below groundwater level, we need to change the reference group to below groundwater level to base the model on.

There is the possibility: "Groundwater_levelbelow:SUM.PAHs..EPA." which will indicate how the taxa 'respond to PAH levels below groundwater level' but the interaction term changes the interpretation greatly. So be carefull with the interpretation.
```{r plot LinDA results, fig.width=config$Differential_Abundance$plotting$fig_width, fig.height=config$Differential_Abundance$plotting$fig_height}
config <- yaml::read_yaml(config_path)
# Define which effect to visualize from Config
var_to_plot <- config$Differential_Abundance$plotting$var_to_plot

log_fold_cutoff_plotting <- "0.5"

colour_positive <- "#17becf"

colour_negative <- "#f8766d"
## dodgerblue4, firebrick4
### 17becf, #f8766d
### DARK : #0a0d36
names(linda_res$output)
#>  names(linda_res$output)
# [1] "time_continuous"                                   "sample_depth_information0.60-0.80"                
#[3] "manganese2_mg_l"                                   "iron2_mg_l"                                       
#[5] "concentration_BTEX_ug_l"                           "phosphorus_total"                                 
#[7] "time_continuous:sample_depth_information0.60-0.80"

var_to_plot <- "CW_number_MF_numberCW_3_MF05"
# Run the plotting script
source(config$scripts$linda_plot)
```


Small chunk of code to test which models are the best 
```{r, }

# Calculate your overall Aitchison (Euclidean on CLR) distance matrix
distance_matrix <- vegdist(t(otu_table(mibi_bmr_clr)), method = "euclidean")

# Model 1: Additive formula (No Interaction)
permanova_additive <- adonis2(distance_matrix ~ time_continuous + sample_depth_information + 
                              electron_donor_acceptor_balance_BTEXN_PAH_acceptors_consumed_relative_to_1D_flow + concentration_BTEX_ug_l + phosphorus_total, 
                              data = linda_metadata, permutations = 999)

print(permanova_additive)
# Model 2: Interaction formula (Time * Depth)
permanova_interaction <- adonis2(distance_matrix ~ time_continuous * sample_depth_information + 
                                 electron_donor_acceptor_balance_BTEXN_PAH_acceptors_consumed_relative_to_1D_flow + concentration_BTEX_ug_l +phosphorus_total, 
                                 data = linda_metadata, permutations = 999)
print(permanova_interaction)


```

# 8.1 Microbial network analysis: Filtering of the phyloseq objectgeneration of the network object.
```{r}
library(NetCoMi)
# =====================================================================
# 1. Selection and filtering of the phyloseq object to use for network selection
# =====================================================================
# =====================================================================
# First: Select the phyloseq object choose the original filtered for interest data table with no prior normalisation. because the function itself allows for multiple normalisation methods internally
# =====================================================================
ps_network_analysis <- ps_work
## 1. Get the names of the first 5 samples
#first_50_samples <- sample_names(ps_network_analysis)[1:162]
## 2. Subset the phyloseq object
#ps_debug <- subset_samples(ps_network_analysis, sample_names(ps_network_analysis) %in% first_50_samples)
#
#
#
#ps_debug # with this the pipeline can be tested/debugged
#otu_debug <- as(otu_table(ps_debug), "matrix")
#
## Agglomerate to genus level
#ps_debug_genus <- tax_glom(ps_debug, taxrank = "genus")
#
## Rename taxonomic table and make genus unique, Because the taxonomic level which you want to use to make the networks, must be unique.
#ps_debug_genus_renamed <- renameTaxa(ps_debug_genus, 
#                                  pat = "<name>", 
#                                  substPat = "<name>_<subst_name>(<subst_R>)",
#                                  numDupli = "genus")
##Save this object for later network construction
#saveRDS(ps_debug_genus_renamed, file = #"Objects_Generated_In_Pipeline/Network_Analysis/ps_network_analysis_Genus_renamed_wetland_samples.rds")
# =====================================================================
# Load in the previously generated network analysis object with renamed genus,...
# =====================================================================
ps_debug_genus_renamed <- readRDS("Objects_Generated_In_Pipeline/Network_Analysis/ps_network_analysis_Genus_renamed_wetland_samples.rds")
# =====================================================================
# 2. Subsetting and filtering
# ====================================================================
ps_debug_genus_renamed_selection <- subset_samples(ps_debug_genus_renamed, timepoint %in% c("T0","T1","T2","T3"))

ps_debug_genus_renamed_selection_2layer_filtering <- subset_samples(ps_debug_genus_renamed_selection, CW_number %in% c("CW_2"))
#Rename it after you do the two layer filtering :
ps_debug_genus_renamed_selection <- ps_debug_genus_renamed_selection_2layer_filtering
#ps_debug_genus_renamed_selection <- subset_samples(ps_debug_genus_renamed, timepoint %in% c("T1", "T2", "T3"))
#ps_object_for_network_analysis <-ps_debug_genus_renamed_selection  # incase you don't want to do further filtering
# =====================================================================
# Further filtering for the network analysis, prevalence 
# =====================================================================
# Keep only genera that are present in at least 25% of your samples 

prevalence_threshold <- 0.25 * nsamples(ps_debug_genus_renamed_selection)
core_genera <- taxa_names(filter_taxa(ps_debug_genus_renamed_selection, function(x) sum(x > 0) >= prevalence_threshold, TRUE))
# Create a temporary phyloseq object with only your widespread taxa, this is 
ps_prevalence_filtered <- prune_taxa(core_genera, ps_debug_genus_renamed_selection)
# here you select the object which will be used for the network analysis
ps_object_for_network_analysis <- ps_prevalence_filtered

```

# Network construction step:
```{r}
# =====================================================================
# =====================================================================
# Network construction step, there are multiple ways of constructing the network: 
# =====================================================================
# =====================================================================
# There are 3 main ways of associating the taxa of interest:
# Correlation, proportionality and conditional dependence
# From each of these measures, there are also different ways of data filtering, zero handling and normalization methods.
# I have included one section for each one of these major association methods to generate the networks, since the choice of association has a great impact on the structure of the final network.
### Specific parameters customisation is mentioned in the 'help' files associated with the function, for some important parameters, short information is included in the current script.
                                       # help("netConstruct")
# An important one which is important for the visualisation: dissFunc: there are multple measures to calculate the dissimilarity between the different taxa which will determine the position of the taxa within the final visualisation. And therefore also the interpretation or intuition of the resulting network. 
    #-**unsigned**: leading to low distance between strongly **associated** taxa (both positively associated as well as negatively associated) 
  #-**signed:** where the distance between the nodes is highest for strongly negative associated taxa. (default)
  #-**SignedPos,** signed distance with setting negative assocations to zero.
  #-There is also TOMdiss: dissimilarity measure available  based on the topological overlap matrix (TOM), instead of looking at the relationships between the taxa in isolation, it evaluates the shared neighborhood context. (if taxa share a lot of the same neighbors, they will be closer to one another)
# =====================================================================
# Correlation based methods
# =====================================================================
# We start off with Correlation, since this is the most intuitive and 'straight forward method:
  ## Correlation-based networks represent co-occurrence patterns (marginal associations) rather than direct biological   interactions. In these networks, if Taxon A and Taxon B both thrive in the same environmental niche (e.g., sharing a specific depth or pH preference), they will show a strong positive correlation. However, this correlation is often a "ghost" or spurious association driven by a shared environmental confounder rather than a direct relationship.

  ## To identify actual direct relationships, conditional dependence methods (such as SPRING) are used. These methods evaluate whether Taxon A and Taxon B remain associated after conditioning on (controlling for) all other taxa in the dataset (Taxon C,D,...). This statistical control actively filters out indirect paths and shared environmental effects.

## Network Construction using SparCC: For our correlation-based analysis, we utilized SparCC (Sparse Correlations for Compositional data). Traditional correlation metrics (like Pearson or Spearman) yield spurious results on microbiome datasets due to their compositional nature (where relative abundances must sum to 1). SparCC is specifically designed to be compositionally aware, utilizing log-ratio transformations to estimate true correlation coefficients without compositionality bias.

### Here normalisation and zero replacement methods are automatically selected so there is no need to specify it 
# The correlation based network construction script: 

network_correlation_based <- netConstruct(ps_object_for_network_analysis,  
                          measure = "sparcc",
                          #normMethod = "clr", This normalisation method is not needed for sparcc
                          filtTax = "highestFreq",
                          group = as.character(phyloseq::sample_data(ps_object_for_network_analysis)$sample_depth_information),
                          taxRank = "genus",
                          filtTaxPar = list( highestFreq = 50),
                          #zeroMethod = "multRepl", This is not needed for sparcc
                          sparsMethod = "threshold",
                          thresh = 0.3,
                          dissFunc = "TOMdiss", # There are different methods for calculating the dissimilarity signed"(default), "unsigned", "signedPos", "TOMdiss"
                          verbose = 3,seed = 123456)

props_SparCC <- netAnalyze(network_correlation_based, 
                          clustMethod = "cluster_fast_greedy")

# =====================================================================
# Conditional dependence based method
# =====================================================================
## Conditional Dependence (Direct Relationships): This approach determines if Taxon A and Taxon B are still connected once we statistically account for the rest of the community (Taxon C). It strips away these indirect, environmentally-driven "ghost" connections to reveal the true backbone of the network.

## We do the network constuction using the SPRING, Unlike traditional correlation metrics that yield spurious results on compositional microbiome data (where relative abundances must sum to 1), SPRING is fully compositionally aware.This allows SPRING to estimate a sparse precision matrix representing true conditional independence, successfully distinguishing direct ecological interactions from indirect background noise.

### Here normalisation and zero replacement methods are automatically selected so there is no need to specify it 
# The conditional dependence based network construction script: 
network_conditional_dependence_based <- netConstruct(ps_object_for_network_analysis,
                           #or the prevalence filtered taxa:ps_prevalence_filtered
                           taxRank = "genus",
                           filtTax = "highestFreq",
                           filtTaxPar = list( highestFreq = 50), #select the number of taxa #filtTaxPar = list(highestFreq = 40)
                           filtSamp = "none",
                           measure = "spring",
                           group = as.character(phyloseq::sample_data(ps_object_for_network_analysis)$sample_depth_information),# Here adjust it to the input object, can be used to generate multiple networks
                           measurePar = list(nlambda=10, 
                                             rep.num=10,
                                             Rmethod = "approx"),
                           dissFunc = "signed", # There are different methods for calculating the dissimilarity signed"(default), "unsigned", "signedPos", "TOMdiss", Signed if you want to look at the direct relationships between the different taxa. TOMdiss, if you want to value the taxon's other connections.
                           verbose = 2,
                           seed = 123456)
library(remotes)
# =====================================================================
# Proportionality based method
# =====================================================================
#############################
## Network Construction using Proportionality (rho)

#We construct the microbial networks using **Proportionality (rho)**. Unlike traditional correlation metrics (such as #Pearson or Spearman) that yield spurious (fake) associations on compositional relative abundance data, the #proportionality measure rho is fully compositionally aware and sub-compositionally coherent. It is a symmetric measure #with values scaled between -1 and +1 (where +1 represents perfect positive proportionality), making it directly #comparable to standard correlation coefficients and suitable for standard network topology algorithms.

#### Connection to Pearson Correlation
#The formula for rho is highly analogous to the classic **Pearson Correlation Coefficient (r)**. 
#Recall that Pearson's correlation scales covariance using the geometric mean of the variances:
#r(Ai, Aj) = cov(Ai, Aj) / square_root([var(Ai)]^2 * [var(Aj)]^2)
#In contrast, proportionality rho scales the relationship using the arithmetic mean of the variances. It is calculated as: rho(log(x_i),log(x_j) = 1 - [ var( log(x_i / x_j) ) / ( var(log(x_i)) + var(log(x_j)) ) ]

#### Breaking Down the Formula Components:
#
#* **log(x_i / x_j)**: Takes the logarithm of the ratio of the two taxa. This is mathematically equivalent to the #difference of their logs: log(x_i) - log(x_j). if this log(x_i)- log(y_i) = a constant value, this means that the ratio's of these taxa are roughly equal overall and, then the variance of this term is low.=>
#* **var( log(x_i / x_j) )**: Calculates the Log-Ratio Variance (VLR). If the ratio between two taxa stays highly consistent across different samples/conditions, this variance will be very close to 0.
#* **var(log(x_i)) + var(log(x_j))**: Represents the sum of the individual variances of the log-transformed data, acting as a scaling/standardising factor. (Note: In practice, to make this compositionally safe, NetCoMi calculates this using Centered #Log-Ratio (CLR) transformed values).
#### The Three Core Scenarios for interpreting the rho scores:
#
##### 1. Perfect Positive Proportionality (rho = 1)
#If Taxon x is always for example exactly 3 times more abundant than Taxon y, their ratio x/y is a constant (3). 
#* The log of a constant is a constant (log(3) is approximately 1.099).
#* The variance of a constant across samples is 0. Plugging this into the numerator of the formula yields:
#rho = 1 - [ 0 / (var(log(x)) + var(log(y))) ] = 1 - 0 = 1

##### 2. Complete Independence (rho = 0)
#If the two genera have absolutely nothing to do with each other, the variance of their difference is simply the sum of their individual variances: 
#var(log(x) - log(y)) = var(log(x)) + var(log(y))
#* Under this condition, the numerator becomes identical to the denominator.Therefore the fraction simplifies to 1, yielding:
#rho = 1 - 1 = 0

##### 3. Perfect Inverse Proportionality (rho = -1)
#If one taxon increases while the other decreases, their covariance is highly negative. The variance of their difference expands because of the algebraic rule:
#var(log(x) - log(y)) = var(log(x)) + var(log(y)) - [ 2 * cov(log(x), log(y)) ]
#* Because covariance is negative, subtracting it becomes addition: - [ 2 * (-covariance) ] = + [ 2 * #absolute_value_of_covariance ]. 
#* This makes the variance of their difference (the numerator) significantly larger than the sum of their individual  variances (the denominator).
#* Due to mathematical boundaries, the numerator can never exceed twice the value of the denominator. Thus, the fraction #is capped at 2:
#rho = 1 - 2 = -1
#* This lower limit of -1 represents a perfect reciprocal relationship. Because the raw data is first CLR-transformed,  even exponential differences between competing bacteria are successfully projected into this symmetric, linear -1 to +1  space.
######
### To achieve this, rho  modifies the log-ratio variance using Centered Log-Ratio (clr) transformed data. This provides a #scaled, robust association metric that is mathematically immune to compositionality bias, allowing us to reconstruct true #co-abundance patterns without the risk of false-positive correlations.

## We do the network constuction using the propr package, which uses clr transformed data
### Here normalisation and zero replacement methods are automatically selected so there is no need to specify it
library(propr)
network_proportionality_based <- netConstruct(ps_object_for_network_analysis, 
                                            taxRank = "genus", 
                                            filtTax = "highestFreq", 
                                            filtTaxPar = list(highestFreq = 50), 
                                            filtSamp = "none", 
                                            measure = "propr", 
                                            group = as.character(phyloseq::sample_data(ps_object_for_network_analysis)$sample_depth_information),
                                            # --- ADDED THESE TWO LINES TO FIX THE ERROR ---
                                            sparsMethod = "threshold",
                                            thresh = 0.6, # Adjust this value (e.g., 0.2 to 0.5) depending on how sparse you want your network
                                            # ----------------------------------------------
                                            dissFunc = "signed", # There are different methods for calculating the dissimilarity signed"(default), "unsigned", "signedPos", "TOMdiss"
                                            verbose = 2, 
                                            seed = 123456)

```
Network selection and analysis
```{r}
# Here you make the selection of which network you are going to analyse:
network_to_analyse <- network_proportionality_based #network_correlation_based, network_conditional_dependence_based or network_proportionality_based
# =====================================================================
# Now we start with the network analysis, there is a lot that can be done with the generated network
# =====================================================================
#Here you can visualize both the association matrix and the similarity matrix
head(network_to_analyse$edgelist1)
plotHeat(network_to_analyse$assoMat1, textUpp = "none", textLow = "none", labCex = 0.7) #labCex, to adjust the label size of the matrix
plotHeat(network_to_analyse$simMat1, textUpp = "none", textLow = "none",labCex = 0.7 )

# In the following function we peform the network analysis:
# There are numerous ways of analysing the data, pls read the original documentation to see all different methods: https://doi.org/10.1093/bib/bbaa290 
## Centrality related: centrLCC = TRUE, very well documented in the help(netAnalyze), best left on true,this makes sure that
## normDeg: default is true, so a normalised version of different centrality values is returned, I keep this out of the customizable parameters, since it is logical and intutive to keep it this way.
## weightDeg, here you choose wheter the strength of the connections plays a role in the connectivity scores, for examples, do you value if a bacterium is connected with 5 small connections the same as 5 big associations? do you value the weight of the connections? or just the number (intuitivly use weightDeg= TRUE)
##########
##centrLCC = FALSE 
#####
props_network_to_analyse <- netAnalyze(network_to_analyse, 
                           centrLCC = TRUE, 
                           weightDeg = TRUE,
                           clustMethod = "cluster_fast_greedy",
                           hubPar = "eigenvector", 
                           hubQuant = 0.95 # quantile for determing hub nodes, set more stringent if you want fewer 'hub' species
                           )
# here you have a large amount of network characteristics 
summary(props_network_to_analyse, numbNodes = 5L)

# =====================================================================
# In this part we want to plot the network, here again there are a million ways of plotting the network, dependent on the manner the edges and node positiontioning is done, this can be done on multiple network traits.
# =====================================================================
plot_network <- plot(props_network_to_analyse, 
          labelScale = FALSE,
          nodeColor = "cluster",  
          nodeSize = "eigenvector", #"eigenvector"
          repulsion = 0.5, #0.8, the lower you go 
          rmSingles = TRUE,
          cexLables = 1.3,
          nodeSizeSpread = 3,
          cexNodes = 2,
          title1 = "Network on genus level with choosen associations", 
          showTitle = TRUE,
          cexTitle = 2.1,
          cexLabels = 1.5)

legend(0.7, 1.1, cex = 1, title = "estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"), 
       bty = "n", horiz = TRUE)

#Snowflake plots 

# Snowflake visualisation 
```{r}
devtools::load_all("scripts/Snowflake")
library(jsonlite)
library(webshot)
Object_For_Snowflake <-ps_work

#generate the microbiome composition object:
# You can define a specific beta divesity metric to later use to colour in the source nodes:unifrac 
MCG_Snowflake<- build_mc_graph(ps_object = Object_For_Snowflake, beta_dist = "wunifrac")
#Differet visualisation methods:
set.seed(12)
# Colour in the source nodes with information with regard to the alfa diversity
forceGraph(graph_object = MCG_Snowflake, source_color = "alphaObserved", target_color = "observedIn", nodeTitle = c("id", "genus", "abundance"), linkStroke = "#DEE2E6")

# Colour in the source nodes with information with regard to the beta diversity ordination (bray curtis in this case)
forceGraph(graph_object = MCG_Snowflake, source_color = "beta", target_color = "observedIn", nodeTitle = c("id", "genus", "abundance"), linkStroke = "#DEE2E6")

temp_mcg <- MCG_Snowflake
  
  
# Colour in the source nodes by setting a specific variable as the grouping varriable
temp_mcg$nodes$group <- temp_mcg_version1$nodes$timepoint

# Make sure to include both the groupingVariable = "group", source_color = "group" arguments, otherwise it won't work

forceGraph(graph_object = temp_mcg, groupingVariable = "group", source_color = "group", target_color = "abundance", nodeTitle = c("group","id", "genus", "abundance"), linkStroke = "#DEE2E6")

###
```
```{r}



```


```{r}
devtools::load_all("scripts/Snowflake")
set.seed(12)
#Object_For_Snowflake <-ps_work
# 1. Build the graph object normally
#MCG_Snowflake<- build_mc_graph(ps_object = Object_For_Snowflake, beta_dist = "wunifrac")

#Add a grouping varriable for plotting
MCG_Snowflake$nodes$group <- MCG_Snowflake$nodes$timepoint
library(htmltools)
# Define your save path
save_path <- "results/Snowflake_Results"
if (!dir.exists(save_path)) dir.create(save_path, recursive = TRUE)
#Different ways of filtering out the 'Target Genera'
        # 1. Filter for genera found in > 1 sample
        multi_sample_nodes <- MCG_Snowflake$nodes[
          !is.na(MCG_Snowflake$nodes$genus) & MCG_Snowflake$nodes$observedIn > 6, 
        ]
        target_genera <- names(sort(table(multi_sample_nodes$genus), decreasing = TRUE))
        # 2. Filter for genera found in a certain abundance cutoff
        abudant_genera_nodes <- MCG_Snowflake$nodes[
          !is.na(MCG_Snowflake$nodes$genus) & MCG_Snowflake$nodes$abundance > 0.001, 
        ]
        target_genera <- names(sort(table(abudant_genera_nodes$genus), decreasing = TRUE))

# 3. Create the Grid HTML 
# We remove the background and shadows to keep the immage clean for editing
grid_items <- lapply(target_genera, function(g_name) {
  highlight_vector <- c("p1","p2","p3","p4","p5", g_name)
  
  p <- forceGraph(
    graph_object = MCG_Snowflake,
    target_color = highlight_vector,groupingVariable = "group", source_color = "group",
    nodeTitle = c("id", "abundance", "genus", "Sample_NAME")
  )
  tags$div(
    style = "display: inline-flex; 
             flex-direction: column; 
             align-items: left; 
             justify-content: center; 
             width: 44%; 
             margin: 1%; 
             vertical-align: top;
             horizonatal-align: left;
             background: white; position: relative;",
    tags$h4(g_name, style = "font-family: sans-serif; text-align: center;align-items: center"),
    p
  )
})

html_grid <- tags$div(
  style = "width: 1800px; background: grey;align-items: center",
  tagList(grid_items)
)

# 4. Save Temp HTML
temp_html <- file.path(save_path, "temp_for_svg.html")
save_html(html_grid, temp_html)

# 5. Export to png
# webshot2 uses Chromium to render the page and can export directly to SVG
webshot(
  url = temp_html,
  file = file.path(save_path, "Snowflake_Final_Figure.png"),
  vwidth = 1750,
  vheight = 1600, delay = 5
)

message("PNG Figure saved to: ", file.path(save_path, "Snowflake_Final_Figure.png"))

```







```
# Plotting with phylum information
```{r}
network_to_analyse <- network_proportionality_based
props_network_to_analyse <- netAnalyze(network_to_analyse, 
                           centrLCC = TRUE, 
                           weightDeg = TRUE,
                           clustMethod = "cluster_fast_greedy",
                           hubPar = "eigenvector", 
                           hubQuant = 0.95 # quantile for determing hub nodes, set more stringent if you want fewer
)

                           # Get phyla names
taxtab_network_analysis <- as(tax_table(ps_object_for_network_analysis), "matrix")
phyla <- as.factor(gsub("p__", "", taxtab_network_analysis[, "phylum"]))
print(phyla)
names(phyla) <- taxtab_network_analysis[, "genus"]

##################################################### When you want to give each phylum a seperate colours ###############
## 1. Convert phyla to a factor to ensure clean levels
#phyla_factor <- as.factor(phyla)
## 2. Count how many unique phyla you actually have
#num_colors <- nlevels(phyla_factor)
## 3. Create a color palette with that exact number (e.g., 52)
#phylcol <- rainbow(num_colors)
## Colors used in the legend should be equally transparent as in the plot
#phylcol_transp <- colToTransp(phylcol, 60)
##Second option: colour them based on the top X phyla:
###################################################### When you just want to give the top X phyla ##############
# 1. Convert phyla to a character vector
phyla_clean <- as.character(phyla)

# 2. Find the top 10 most common phyla
top_phyla <- names(sort(table(phyla_clean), decreasing = TRUE)[1:10])

# 3. Replace any phylum not in the top 10 with "Other"
phyla_clean[!(phyla_clean %in% top_phyla)] <- "Other"

# 4. Convert it back to a factor
phyla_grouped <- as.factor(phyla_clean)

# ==================== THE FIX ====================
# Assign the original taxon names to your new grouped factor vector
names(phyla_grouped) <- names(phyla)
# =================================================

# 5. Define your 11 colors
#The bright version
#phylcol <- c(
#  "cyan", "blue3", "red", "lawngreen", "yellow", "deeppink", 
#  "darkorange", "purple", "saddlebrown", "turquoise4", "gray60"
#)

phylcol <- c(
  "#00A896", # 1. Bright Teal (distinct from blues and greens)
  "#023E8A", # 2. Deep Royal Blue (strong, dark anchor)
  "#E63946", # 3. Vibrant Coral-Red (pops instantly)
  "#40916C", # 4. Emerald Green (rich and organic)
  "#FFB703", # 5. Bright Amber/Gold (highly visible, warm)
  "#D81159", # 6. Vivid Raspberry (distinct from red and purple)
  "#F77F00", # 7. Intense Ochre Orange (strong warmth)
  "#7209B7", # 8. Deep Electric Purple (royal and distinct)
  "#6F4E37", # 9. Rich Coffee Brown (neutral anchor)
  "#00F5D4", # 10. Neon Mint (your high-pop accent color)
  "#4A5568"  # 11. Slate Gray (clean neutral for minor phyla)
)
phylcol_transp <- colToTransp(phylcol, 30)
# ==================== DRAW THE PLOT ====================
plot(props_network_to_analyse,
     layout = "spring",
     repulsion = 0.4,
     shortenLabels = "none",
     charToRm = "g__",
     labelScale = FALSE,
     rmSingles = TRUE,
     nodeSize = "clr", # eigenvector, clr
     nodeSizeSpread = 4,
     nodeColor = "feature", 
     featVecCol = phyla_grouped, 
     colorVec =  phylcol,
     posCol = "#2A6F97", # "darkturquoise" or "#2A6F97"
     negCol = "#FF2400", # "orange"
     edgeTranspLow = 15,
     edgeTranspHigh = 40,
     cexNodes = 1.3,
     cexLabels = 1.3,
     cexHubLabels = 1.8,
     title1 = "Network on genus level with Pearson correlations", 
     showTitle = TRUE,
     cexTitle = 2.3)
# color them on cluster
plot(props_network_to_analyse,
     layout = "spring",
     repulsion = 0.4,
     shortenLabels = "none",
     charToRm = "g__",
     labelScale = FALSE,
     rmSingles = TRUE,
     nodeSize = "clr", # eigenvector, clr
     nodeSizeSpread = 4,
     nodeColor = "cluster",
     posCol = "#2A6F97", # "darkturquoise" or "#2A6F97"
     negCol = "#FF2400", # "orange"
     edgeTranspLow = 15,
     edgeTranspHigh = 40,
     cexNodes = 1.3,
     cexLabels = 1.3,
     cexHubLabels = 1.8,
     title1 = "Network on genus level with Pearson correlations", 
     showTitle = TRUE,
     cexTitle = 2.3)
# ==================== DRAW THE LEGENDS ====================

# Allow plotting slightly into the margins
par(xpd = TRUE) 

# Phylum Legend: Nudged left (-1.3), up (1.4), and vertically compressed (y.intersp = 0.7)
legend(-1.3, 1.4, 
       cex = 0.8,                 # Slightly smaller text to keep it tidy
       pt.cex = 1.2, 
       title = "Phylum:",        
       legend = levels(phyla_grouped), 
       col = phylcol_transp, 
       bty = "n", 
       pch = 16,
       y.intersp = 0.7)           # Squeezes the 11 items vertically so they don't hit the nodes!

# Association Legend: Nudged right (0.6) and up (1.4) to match the left side
legend(0.6, 1.4, 
       cex = 0.8,                 # Kept at 0.8 to match the Phylum legend scale
       title = "Estimated association:",
       legend = c("+","-"), 
       lty = 1, 
       lwd = 3, 
       col = c("darkturquoise","orange"), 
       bty = "n", 
       horiz = TRUE,
       x.intersp = 0.8)           # Squeezes the horizontal spacing slightly

# Reset plotting boundaries back to default
par(xpd = FALSE)

```

Save the plot in a map
```{r}
# Now we save the plots: 
pdf("Objects_Generated_In_Pipeline/Network_Analysis/netcomi_spring.pdf", width = 20, height = 10)

plot(props_network_to_analyse, 
          labelScale = FALSE,
          nodeColor = "cluster", 
          nodeSize = "eigenvector",
          repulsion = 0.8,
          rmSingles = TRUE,
          cexLables = 1.5,
          nodeSizeSpread = 3,
          cexNodes = 2,
          title1 = "Network on genus level with SPRING associations", 
          showTitle = TRUE,
          cexTitle = 2.1,
          cexLabels = 1.5)
dev.off()


```





# 8.2 Microbial network analysis: analyzing the generated network
```{r}

#help("netAnalyze")


##########

#summary.microNetProps
summary(props_spring, numbNodes = 5L)
p <- plot(props_spring, 
          labelScale = FALSE,
          nodeColor = "cluster", 
          nodeSize = "eigenvector",
          repulsion = 0.8,
          rmSingles = TRUE,
          cexLables = 1.3,
          nodeSizeSpread = 3,
          cexNodes = 2,
          title1 = "Network on genus level with SPRING associations", 
          showTitle = TRUE,
          cexTitle = 2.1,
          cexLabels = 1.5)

legend(0.7, 1.1, cex = 1, title = "estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"), 
       bty = "n", horiz = TRUE)


# Now we save the plots: pls write a script to do this: 
pdf("Objects_Generated_In_Pipeline/Network_Analysis/netcomi_spring.pdf", width = 20, height = 10)

plot(props_spring, 
          labelScale = FALSE,
          nodeColor = "cluster", 
          nodeSize = "eigenvector",
          repulsion = 0.8,
          rmSingles = TRUE,
          cexLables = 1.5,
          nodeSizeSpread = 3,
          cexNodes = 2,
          title1 = "Network on genus level with SPRING associations", 
          showTitle = TRUE,
          cexTitle = 2.1,
          cexLabels = 1.5)

dev.off()

# 

# For Gephi, we have to generate an edge list with IDs.
# The corresponding labels (and also further node features) are stored as node list.

# Create edge object from the edge list exported by netConstruct()
edges <- dplyr::select(net_spring$edgelist1, v1, v2)

# Add Source and Target variables (as IDs)
edges$Source <- as.numeric(factor(edges$v1))
edges$Target <- as.numeric(factor(edges$v2))
edges$Type <- "Undirected"
edges$Weight <- net_spring$edgelist1$adja

nodes <- unique(edges[,c('v1','Source')])
colnames(nodes) <- c("Label", "Id")

# Add category with clusters (can be used as node colors in Gephi) Here you can select some node characteristics to take out to make a better plot in gephi
nodes$Category <- props_spring$clustering$clust1[nodes$Label]

edges <- dplyr::select(edges, Source, Target, Type, Weight)

write.csv(nodes, file = "Objects_Generated_In_Pipeline/Network_Analysis/Exporting_To_Gephi/nodes.csv", row.names = FALSE)
write.csv(edges, file = "Objects_Generated_In_Pipeline/Network_Analysis/Exporting_To_Gephi/edges.csv", row.names = FALSE)

#Network with Pearson correlation as association measure 
net_pears <- netConstruct(ps_prevalence_filtered,  
                          measure = "pearson",
                          normMethod = "clr",
                          filtTax = "highestFreq",
                          group = as.character(phyloseq::sample_data(ps_prevalence_filtered)$sample_depth_information),
                          taxRank = "genus",
                          filtTaxPar = list( highestFreq = 50),
                          zeroMethod = "multRepl",
                          sparsMethod = "threshold",
                          thresh = 0.3,
                          verbose = 3)

props_pears <- netAnalyze(net_pears, 
                          clustMethod = "cluster_fast_greedy")


plot(props_pears, 
     nodeColor = "cluster", 
     nodeSize = "eigenvector",
     repulsion = 0.8,
     rmSingles = TRUE,
     labelScale = FALSE,
     cexLables = 1.3,
     nodeSizeSpread = 3,
     cexNodes = 2,
     title1 = "Network on genus level with Pearson correlations", 
     showTitle = TRUE,
     cexTitle = 2.3)

legend(0.7, 1.1, cex = 2.2, title = "estimated correlation:", 
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"), 
       bty = "n", horiz = TRUE)


pdf("Objects_Generated_In_Pipeline/Network_Analysis/netcomi_pearson.pdf", width = 20, height = 10)

plot(props_pears, 
     nodeColor = "cluster", 
     nodeSize = "eigenvector",
     repulsion = 0.8,
     rmSingles = TRUE,
     labelScale = FALSE,
     cexLables = 1.3,
     nodeSizeSpread = 3,
     cexNodes = 2,
     title1 = "Network on genus level with Pearson correlations", 
     showTitle = TRUE,
     edgeInvisFilter = "threshold", #if you want to filter out some of the edges, make them invisible
     edgeInvisPar = 0.5,
     cexTitle = 2.3)

dev.off()


```
Correlation based networks: SParCC
```{r}



plot(props_SparCC, 
     nodeColor = "cluster", 
     nodeSize = "eigenvector",
     repulsion = 0.8,
     rmSingles = TRUE,
     labelScale = FALSE,
     cexLables = 1.3,
     nodeSizeSpread = 3,
     cexNodes = 2,
     title1 = "Network on genus level with Pearson correlations", 
     showTitle = TRUE,
     cexTitle = 2.3)

legend(0.7, 1.1, cex = 2.2, title = "estimated correlation:", 
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"), 
       bty = "n", horiz = TRUE)


pdf("Objects_Generated_In_Pipeline/Network_Analysis/netcomi_SparCC.pdf", width = 20, height = 10)

plot(props_SparCC, 
     nodeColor = "cluster", 
     nodeSize = "eigenvector",
     repulsion = 0.8,
     rmSingles = TRUE,
     labelScale = FALSE,
     cexLables = 1.3,
     nodeSizeSpread = 3,
     cexNodes = 2,
     title1 = "Network on genus level with SParCC correlations", 
     showTitle = TRUE,
     edgeInvisFilter = "threshold", #if you want to filter out some of the edges, make them invisible
     edgeInvisPar = 0.5,
     cexTitle = 2.3)

dev.off()


```






#TAX4FUN Section
## 7.1 Functional assignment using tax4fun
### This chunk of code, is used for doing the functional assignment, which is ran in the 09_1_Tax4Fun_object_generation.R script. Run this one time per dataset, since this takes a lot of time 
Chunk of code to for tax4fun_object_generation, this step takes time so run once per dataset and then load in the objects during the analysis
```{r, results= 'hide'}
config <- yaml::read_yaml(config_path)
#This is what we need to specify for the analyis:
# modify the following 8 lines as needed
query_otu_seq   = 'tax4fun2/mibi_pipeline_otus.fasta'       # input OTU sequence file, this is generated in the .R script based on the ps_work phyloseq object
query_otu_table = 'tax4fun2/mibi_pipeline_otu_table.txt'    # input OTU table, this is generated in the .R script
pwd_op_folder   = 'tax4fun2/Tax4Fun2_output_folder'      # output directory
pwd_ref_data    = 'tax4fun2/Tax4Fun2_ReferenceData_v2'   # path to the default database (i.e., Tax4Fun2_ReferenceData_v2), need to be decompressed before use, this was normally done during the instalation process.
norm_by_cn      = TRUE                          # normalize_by_copy_number (TRUE or FALSE)
norm_path       = TRUE                          # normalize_pathways (TRUE or FALSE)
iden            = 0.97                          # min_identity_to_reference, please modify as needed
num_of_threads  = 2                             # number of CPU cores to use, please modify as needed
#when you want to run the analysis, uncheck this #
#source(config$scripts$tax4fun_object_generation)
```
## 7.2 Filtering out the KO's of interest for further analysis
### This chunk of code, you can take out the information generated by the tax4fun function. And select in which collumns you are interested in, this filters the Kegg orthology pathways you are interested in, which in the end can be linked as data to your phyloseq data.
```{r, results= 'hide'}
config <- yaml::read_yaml(config_path)
# 1. Set paths to your generated output files
pathway_file <- "tax4fun2/Tax4Fun2_output_folder/pathway_prediction.txt"
ko_file      <- "tax4fun2/Tax4Fun2_output_folder/functional_prediction.txt"

# 1. Define the exact categories you WANT to keep in this case we use level_2_to_filter
# An overview of these categories can be seen by manually run the code within the 09_2_Tax4Fun_object_filtering.R script 

# Select on which level you want to filter the tax4fun object:
filter_on_KO_level <- 'level_2' # level_2 or level_3

wanted_categories_level_3 <- c("Metabolism", "Genetic Information Processing", "Environmental Information Processing")
#wanted_categories_level_2 <- c("Xenobiotics biodegradation and metabolism","Biosynthesis of other secondary metabolites","Energy metabolism","Metabolism of cofactors and vitamins","Lipid metabolism","Carbohydrate metabolism","Membrane transport","Excretory system")
wanted_categories_level_2 <- c(
  "Xenobiotics biodegradation and metabolism", # <--- HERE IS YOUR PAH/BIOREMEDIATION DEGRADATION!
  "Energy metabolism",                         # <--- HERE IS YOUR METHANE/SULFUR/NITROGEN ANAEROBIC RESPIRATION!
  "Environmental adaptation",                  # Stress responses to anaerobic/toxic conditions
  "Cellular community - prokaryotes"           # Biofilm formation in the wetland substrate
)
#run the filtering script
source(config$scripts$tax4fun_object_filtering)
```
## 7.3 Data object generation for analysis and visualization of the filtered data table
Functional assignment using tax4funNow we have filtered out the pathways of interest, now we can go to data analysis and plotting. 
```{r}
# now we will create a table which can be used for further data visualisation and plotting, and could be added to the metadata already present:
# 1. Isolate biological samples from hierarchical character rows
sample_rows  <- setdiff(rownames(pathway_filtered), c("level1", "level2", "level3"))
# HERE you choose the labels which will be assinged to the specific KOs
#KO_names_for_plotting <- as.character(filtered_pathway_transposed["level2", ])
KO_names_for_plotting <- as.character(pathway_filtered["level1", ])

# 2. Extract and strictly force the numbers to be numeric (Matrix conversion trick)
raw_numeric_part <- pathway_filtered[sample_rows, ]
clean_numeric_matrix <- as.data.frame(matrix(
  as.numeric(as.matrix(raw_numeric_part)), 
  nrow = nrow(raw_numeric_part), 
  ncol = ncol(raw_numeric_part)
))
rownames(clean_numeric_matrix) <- rownames(raw_numeric_part)
colnames(clean_numeric_matrix) <- colnames(raw_numeric_part)

# 3. Transpose the pure numbers and attach their Level 1 grouping categories
transposed_df <- as.data.frame(t(clean_numeric_matrix))
transposed_df$Level1_Category <- KO_names_for_plotting


# Now that we have the data we want, we are going to put it into a format which can be added to the metadata file by copy pasting it
config <- yaml::read_yaml(config_path)
# generate a functional KO table which can be copy pasted into the metadata
source(config$scripts$tax4fun_object_metadata_generation)
#Now you can add this table to your microbiome metadataset for additional analysis
head(metadata_sorted)
```
#Now we can do a preliminary data analysis on the genes of interest found in our sample
```{r}
tax4fun_pathway_data <- metadata_sorted

tax4fun_pathway_data_selection <- tax4fun_pathway_data[, c(105:136) ] 
tax4fun_pathway_data_selection_matrix <- as.matrix(tax4fun_pathway_data_selection)
# This can be different based on the object so check that you have all gene information you are interested in.

#now we impute the zero's so we don't have the zero problem. we use 
tax4fun_pathway_data_selection_matrix_bmr <- zCompositions::cmultRepl(tax4fun_pathway_data_selection_matrix, method = "GBM", output = "p-counts", label = 0, z.delete = 1, z.warning = 1)
# if this gives an error this can mean that there are no zero's observed
tax4fun_pathway_data_selection_matrix_bmr <-  tax4fun_pathway_data_selection_matrix
#now we have handeled the zero problem, so we can move on to the CLR transformation
tax4fun_pathway_data_selection_matrix_bmr_clr    <- microbiome::transform(tax4fun_pathway_data_selection_matrix_bmr, transform = "clr")
# 4. Convert the CLR matrix back to a data frame and pimp the column names
clr_df <- as.data.frame(tax4fun_pathway_data_selection_matrix_bmr_clr)
colnames(clr_df) <- paste0(colnames(clr_df), "_CLR")
# 5. Bind the CLR columns directly onto the back of your original object
# Since the row order has not changed, a direct cbind is perfectly safe
extended_metadata <- cbind(tax4fun_pathway_data, clr_df)

#save the CLR transformed file, raw as an RDS object for potential functional analysis
saveRDS(tax4fun_pathway_data_selection_matrix_bmr_clr, file = "results/all_wetlands/tax4fun_pathway_data_selection_matrix_bmr_clr.rds")
# 5. Write the tab-separated file safely to copy paste behind the metadata with the CLR transformed version included
write.table(extended_metadata, 
            sep = "\t", dec = ",",
            file = "results/all_wetlands/Extended_metadata_plus_CLR.tsv", 
            row.names = FALSE,
            quote = TRUE)

# To run a PCA on these results: see script 09_Run_PCA_on_functional_traits.R 
```



