---
title: "Rmarkdown_16S_Ananlysis_mibibio"
author: "Simon Vandersanden"
date: "2026-03-05"
output:
  html_document: default
  pdf_document: default
params:
  config_file: "configs/Config_Griftpark_Soilcore.yaml"
---

```{r setup, include=FALSE}
#working with a configuration file, to make the workflow more flexible.
library(yaml)
# SAFETY NET: If params doesn't exist (manual run), manually point to the file
if (!exists("params") || is.null(params$config_file)) {
  config_path <- "configs/Config_Griftpark_Soilcore.yaml"
} else {
  config_path <- params$config_file
}
config <- yaml::read_yaml(config_path)
```


## Loading in the necessary libraries for basic handeling

```{r}
library(phyloseq)
library(dplyr)
library(picante)    # For Phylogenetic Diversity
library(microbiome) # For Dominance metrics
library(vegan)      # For Rarefaction curves
library(zCompositions)
library(ggplot2)
library(ggrepel)
library(MicrobiomeStat)
library(scales)
library(paletteer)
library(stringr)
library(yaml)
```

## 1:Loading in and inspecting the different parts of the phyloseq object
Here you load in you microbiome data, there is a part where sample data can be linked to the object, so you can quickly add in new sample data for analysis and plotting.


```{r,include=TRUE, results='hide'}
# --- 1. LOAD PHYLOSEQ OBJECT ---
# This uses the path defined in the YAML file.
mibi_phylo_object <- readRDS(config$paths$phyloseq_rds) 
# Extract tables for inspection
mibi_otu_table  <- as(otu_table(mibi_phylo_object), "matrix")
mibi_tax_table  <- as(tax_table(mibi_phylo_object), "matrix")
mibi_meta_table <- as(sample_data(mibi_phylo_object), "data.frame")

# --- 2. CLEAN TAXONOMY ---
# Instead of "Unclassified", we use the variable from your config
uncl_label <- config$analysis_params$unclassified_label

mibi_tax_table[is.na(mibi_tax_table)] <- uncl_label
mibi_tax_table[mibi_tax_table == ""]  <- uncl_label

# Update the phyloseq object with cleaned taxonomy
tax_table(mibi_phylo_object) <- tax_table(mibi_tax_table)

# --- 3. UPDATE METADATA ---
# We replace the filename, the separator (\t), and the decimal (,) with config values
mibi_meta_table_new <- read.table(
  file = config$paths$metadata_txt, 
  header = TRUE, 
  sep = config$analysis_params$metadata_sep, 
  row.names = 1, 
  stringsAsFactors = FALSE, 
  dec = config$analysis_params$metadata_dec
)

# Safety check
stopifnot(all(sample_names(mibi_phylo_object) %in% rownames(mibi_meta_table_new)))

# Link the new metadata to the object
sample_data(mibi_phylo_object) <- mibi_meta_table_new
```

## 2.1:Data Pre-processing: Filtering out all non-samples and checking rarefaction curves
Normally some pre-filtering has been done on the reads before they were put into the phyloseq object: the sequencing depth threshhold, the decontamination, and some taxonomic filtering of certain undesired taxa such as Eukaryota, Mitochondria and Chloroplast sequences.
### Here we can remove some unecessary samples from the dataset.
### There is also the possibility to plot the rarefaction curves, to see wheter the sequencing depth is sufficiënt.

```{r,include=TRUE}
# It is possible that that some blancs and mock samples are still present in the samples, these should be filtered out because we will only filter using all the 'real samples'
# Filter out unwanted samples using the list from the config
samples_to_remove <- config$filtering$samples_to_remove

mibi_phylo_object <- prune_samples(
  !(sample_names(mibi_phylo_object) %in% samples_to_remove),
  mibi_phylo_object
)
# Run the rarefaction script using the path from config
source(config$scripts$rarefaction)
```
## 2.1:Data Pre-processing: Filtering out based on prevalence and abundance

Often prevalence or abundance filtering is done to remove uncertain features which can be artifacts created. or remove rare sequences, which might not be of interest to us. This is highly customisable so also all features can be kept however often for clarity, features are filtered out.
Here you can filter on prevalence and abundance, you can also select either prevalence or either abundance.

```{r}
# --- Filtering parameters: PULLED FROM CONFIG ---
prev_filter_threshold  <- config$filtering$prev_filter_threshold
abund_filter_threshold <- config$filtering$abund_filter_threshold

# Logic Switches
# We convert the YAML boolean (true/false) to your script's expected 1/0
prev_filter  <- as.numeric(config$filtering$apply_prevalence)
abund_filter <- as.numeric(config$filtering$apply_abundance)

# Run the background script
# Note: Ensure these scripts use the variables defined above!
source(config$scripts$filtering)

# Resulting object 'mibi_phylo_object_filt' is now ready
```

## 2.1:Data Pre-processing: Filtering based on variance OPTIONAL, !!!This is not working yet!!!Also this should not be used for alpha or beta diversity, However it might be usefull before differential abundance analysis
```{r}
# --- Variance parameters: PULLED FROM CONFIG ---
variance_cutoff <- config$variance_filtering$cutoff
variance_filter_switch <- as.numeric(config$variance_filtering$apply_filter)

# Run the background script
# Note: This script uses mibi_phylo_object_filt
source(config$scripts$variance)

# The resulting object 'mibi_phylo_object_filt' is now updated
#This script starts with mibi_phylo_object_filt and updates it with this filtering 
# The resulting object 'mibi_phylo_object_filt' is now updated
```

## 2.1:Data Pre-processing: Sample removal or subset selection
If you have a large dataset or are interested in specific samples, here you can select which samples or groups of samples you want to take a closer look at.
### The script knows whether to subset or keep the full dataset based on the config file
```{r}
if (config$subsetting$use_subset) {
  
  target_group <- config$subsetting$target_value
  target_col   <- config$subsetting$column_name
  
  # Professional trick: use 'get' to find the column name dynamically from the string in YAML
  mibi_phylo_object_subset <- subset_samples(mibi_phylo_object_filt, 
                                             get(target_col) == target_group)
  
  # Remove taxa with zero counts in this specific subset
  mibi_phylo_object_subset <- prune_taxa(taxa_sums(mibi_phylo_object_subset) > 0, 
                                         mibi_phylo_object_subset)
  
  # Reporting
  message(paste("Subset selected:", target_group))
  message(paste("Samples remaining:", nsamples(mibi_phylo_object_subset)))
  message(paste("Taxa remaining after pruning zeros:", ntaxa(mibi_phylo_object_subset)))
  
} else {
  message("No subsetting applied. Using full filtered dataset.")
  mibi_phylo_object_subset <- mibi_phylo_object_filt
}
```
## 2.2 Object selection: selecting the object for the normalisation, and further data analysis.
```{r}
# Define which object flows into the analysis scripts.
# We use the 'subset' object, which is either the actual subset 
# OR the full dataset (if no subsetting was applied in the previous chunk).
ps_work <- mibi_phylo_object_subset
# --- Reporting: Final Object Overview ---
# 1. Get the total counts
n_samples <- nsamples(ps_work)
n_taxa    <- ntaxa(ps_work)
# 2. Get the names and collapse them into one string
sample_list_string <- paste(sample_names(ps_work), collapse = ", ")

# 3. Summary of the samples used for downstream analysis
message("--------------------------------------------------")
message(paste("PROJECT:", config$project_metadata$name))
message(paste("ANALYSIS TARGET:", if(config$subsetting$use_subset) config$subsetting$target_value else "Full Dataset"))
message(paste("TOTAL SAMPLES:", n_samples))
message(paste("TOTAL TAXA:", n_taxa))
message("SAMPLES INCLUDED:")
message(sample_list_string)
message("--------------------------------------------------")

```

# 2.2:Data Pre-processing: Normalisation & The "Zero Problem"

Microbiome data is **compositional**. Because total read counts are an artifact of sequencer depth, they represent "parts of a whole" rather than absolute concentrations. If one taxon blooms, others appear to decrease simply because they occupy a smaller percentage of the total pool.

### 3.1 Handling Sparsity (Zero Imputation)

Log-ratio transformations (like CLR) cannot handle zeros because $\log(0)$ is undefined. We address this using two distinct approaches before transformation:

-   **Simple Pseudo-count (+1):** Adds a constant shift of 1 to every cell. It is fast and robust, though it can slightly bias ratios in extremely sparse datasets.
-   **Bayesian-Multiplicative Replacement (BMR):** Uses the `zCompositions` package to impute zeros based on a Dirichlet prior. This is more mathematically rigorous as it preserves the ratio structure of the data.
    -   **Note:** BMR requires careful handling of "singletons." We have adjusted the `z.delete` and `z.warning` thresholds to ensure samples with limited diversity are not dropped during the imputation process.
-   **Matrix Completion:** *Planned for future implementation.*

------------------------------------------------------------------------
### 3.2 Executed Transformations
The following background script processes the chosen `ps_work` object and generates the normalized data structures, using different methods.
```{r}
# Requires: ps_work (defined above)
# This script will generate: mibi_tss, mibi_pseudo_clr, mibi_bmr_clr

source(config$scripts$normalization)

# Quick check to ensure the objects were created
if(exists("mibi_tss")) {
  message("Normalization complete: mibi_tss, mibi_pseudo_clr, and mibi_bmr_clr are ready.")
}
```

###The script generates three primary objects. Choosing the right one depends on your specific analysis:

-mibi_tss Total Sum Scaling Visualization: Relative Abundance (%) used for stacked bar plots.

-mibi_pseudo_clr Pseudo-CLR (+1) Statistics: Fast transformation for PCoA and basic linear modeling.

-mibi_bmr_clr BMR-CLR (Bayesian) Advanced Stats: High-fidelity modeling and differential abundance.

# Community visualisation: barplots
This chunk is designed to plot relative abundances of you microbial community data, this takes the microbial community and you can select at which level the plotting is done, the labeling is two levels to ensure comprehensiveness, and detail (higher phylogenetic level and lower for the detail). The main color is determined by the high level taxonomy, variations of this main color are used in the lower levels. Several choices can be made, a relative abundance threshold level, which is a cutoff level for relative abundance, if a taxon does not meet this threshold i.e. all taxa with relative abundance below this threshold are put into 'Others'
```{r plot abundance, fig.width=12, fig.height=8}
# --- SETUP ---
source(config$scripts$barplot_main)
# --- 1. INPUT FROM CONFIG ---
v_glob     <- config$Barplot_specific$global_barplots
pal_choice <- config$Barplot_specific$palette_choice

# --- 2. PREPARATION ---
data_for_plot <- prepare_abundance_data(
  ps         = ps_work,
  high_level = v_glob$high_level,
  low_level  = v_glob$low_level, 
  facet_var  = v_glob$facet_var, 
  threshold  = v_glob$threshold 
)
# This prepared the colours, so custom colours are assigned to the created groups depending on the taxonomic specification.
my_colors <- get_shaded_palette(data_for_plot, pal_choice)

# --- 3. DYNAMIC PLOTTING ---
# This is a complex plotting command that is constructed to take into acount the previously specified parameters, high and low taxonomic levels, relative abundance threshold cutoff, how to group the samples for plotting. So this information is also found on the finally generated plot
Relative_Abundance_Plot <- ggplot(data_for_plot, aes(x = !!sym(v_glob$facet_var), y = Abundance, fill = Hierarchical_Label)) +
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.05) +
  scale_fill_manual(values = my_colors) +
  labs(
    title = paste("Relative Abundance:", str_to_title(v_glob$high_level), "-", str_to_title(v_glob$low_level)),
    subtitle = paste("Filtered at", v_glob$threshold, "% threshold | Grouped by", v_glob$facet_var),
    x = stringr::str_replace_all(v_glob$facet_var, "_", " "),
    y = "Relative Abundance (%)",
    fill = paste(str_to_title(v_glob$high_level), "-", str_to_title(v_glob$low_level))
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.text = element_text(size = 9),
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold")
  ) +
  guides(fill = guide_legend(ncol = 2))

Relative_Abundance_Plot

# Save output using config values for naming
ggsave(paste0("results/Relative_Abundance_Plot_", v_glob$high_level, "_", v_glob$low_level, ".png"), 
       plot = Relative_Abundance_Plot, width = v_glob$fig_width, height = v_glob$fig_height, dpi = v_glob$dpi)
```

## Community visualisation, detailed group analysis.
This chunk of code is designed to plot relative abundances of you microbial community data. But here you can look within specific group and get the lower level taxonomic distribution so you can gain insights into specific important groups within you data.The labeling is two levels to ensure comprehensiveness, and detail (higher phylogenetic level and lower for the detail). The main color is determined by the high level taxonomy, variations of this main color are used in the lower levels. Several choices can be made, a relative abundance threshold level, which is a cutoff level for relative abundance, if a taxon does not meet this threshold i.e. all taxa with relative abundance belower this threshold are put into 'Others' 
```{r plot specific group abundance, fig.width=12, fig.height=8}
source(config$scripts$barplot_focus)
# --- 1. DEFINE SPECIFICS FROM CONFIG ---
v_foc      <- config$Barplot_specific$focus_barplot
pal_choice <- config$Barplot_specific$palette_choice

# From this specifically selected group there is the possibility to plot the lower taxonomic levels within it, similar logic to a normal relative abundance plot, but now the relative abundance within this selected taxonomic group is plotted.

# --- 2. PREPARATION ---
data_subset <- prepare_specific_abundance_data(
  ps         = ps_work,
  rank_name  = v_foc$target_rank,
  rank_value = v_foc$target_val,
  high_level = v_foc$high_level,
  low_level  = v_foc$low_level,
  facet_var  = v_foc$facet_var,
  threshold  = v_foc$threshold
)

# This prepared the colors, so custom colors are assigned to the created groups depending on the taxonomic specification.
my_colors_sub <- get_shaded_palette(data_subset, pal_choice)# Uses the shared function from: 08_barchart_visualisation.R

# --- 3. DYNAMIC PLOTTING ---
# This is a complex plotting command that is constructed to take into acount the previously specified parameters, high and low taxonomic levels, relative abundance threshold cutoff, how to group the samples for plotting. So this information is also found on the finally generated plot
Specific_Group_Relative_Abundance_Plot <- ggplot(data_subset, aes(x = !!sym(v_foc$facet_var), y = Abundance, fill = Hierarchical_Label)) +
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.05) +
  scale_fill_manual(values = my_colors_sub) +
  labs(
    title = paste("Focus:", v_foc$target_val, "(", str_to_title(v_foc$high_level), "-", str_to_title(v_foc$low_level), ")"),
    subtitle = paste("Filtered at", v_foc$threshold, "% internal abundance | Grouped by", v_foc$facet_var),
    x = stringr::str_replace_all(v_foc$facet_var, "_", " "),
    y = paste("Relative Abundance within", v_foc$target_val, "(%)"),
    fill = paste(str_to_title(v_foc$high_level), "-", str_to_title(v_foc$low_level))
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.text = element_text(size = 8),
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold", color = "darkblue")
  ) +
  guides(fill = guide_legend(ncol = 2))

Specific_Group_Relative_Abundance_Plot

ggsave(paste0("results/Specific_Group_Relative_Abundance_Plot_", v_foc$target_val, ".png"), 
       plot = Specific_Group_Relative_Abundance_Plot, width = v_foc$fig_width, height = v_foc$fig_height, dpi = v_foc$dpi)
```

## 4.0 Alpha Diversity: Community Trait Analysis

Alpha diversity describes the complexity of a microbial community within a single sample. Rather than looking at "who is there," we look at the mathematical properties of the population structure. We categorize these traits into four pillars:

-   **Richness:** The total number of unique taxa observed (how many species?).
-   **Dominance:** Measures if the community is "top-heavy" with one or two abundant taxa.
-   **Evenness:** How equally the individuals are distributed among the species.
-   **Information:** Mathematical indices (like Shannon) that balance both richness and evenness.

> **Important:** Alpha diversity calculations are sensitive to normalization. We use the filtered, non-transformed object (`mibi_phylo_object_filt`) to ensure we are counting physical reads, not mathematical ratios.

```{r plot different alpha diversity measures, fig.width=12, fig.height=8}
# --- 1. SETUP & INPUT ---
# Use the ps work object
alfa_diversity_phyloseq_object <- ps_work

# Pull traits from Config
selected_richness    <- config$Alpha_Diversity$traits$richness
selected_dominance   <- config$Alpha_Diversity$traits$dominance
selected_evenness    <- config$Alpha_Diversity$traits$evenness
selected_information <- config$Alpha_Diversity$traits$information

# --- 2. RUN CALCULATION SCRIPT ---
# This script uses the variables above to generate 'plot_metrics_all'
source(config$scripts$alpha_calc)

# --- 3. PLOTTING CONFIGURATION ---
plot_x_axis   <- config$Alpha_Diversity$plotting$x_axis
plot_color_by <- config$Alpha_Diversity$plotting$color_by
plot_cols     <- config$Alpha_Diversity$plotting$grid_cols

# Logic to handle 'all' vs specific metrics
if (config$Alpha_Diversity$plotting$selected_metrics == "all") {
  plot_metrics <- plot_metrics_all
} else {
  plot_metrics <- config$Alpha_Diversity$plotting$selected_metrics
}

# --- 4. RUN PLOTTING SCRIPT ---
source(config$scripts$alpha_plot)

# Display and Save
plot(alpha_boxplot)
#We visualize the distribution of community traits across our experimental groups. Note that the Y-axis scales are "free" because the units for Richness (counts) and Dominance (0-1) are fundamentally different.  
ggsave(
  filename = "results/alpha_diversity_plot.png", 
  plot = alpha_boxplot, 
  width = config$Alpha_Diversity$plotting$fig_width, 
  height = config$Alpha_Diversity$plotting$fig_height, 
  dpi = config$Alpha_Diversity$plotting$dpi
)
```

## 5.0 Beta Diversity: Community Dissimilarity

Beta diversity quantifies the "distance" or "dissimilarity" between pairs of samples. To provide a non-redundant overview, we categorize these metrics based on whether they consider abundance (quantitative) vs. presence (qualitative), and phylogeny (evolutionary relatedness).

-   **Aitchison:** Euclidean distance on CLR-transformed data. The gold standard for compositional data.
-   **Bray-Curtis:** Non-phylogenetic; accounts for abundance.
-   **Jaccard:** Non-phylogenetic; accounts only for presence/absence.
-   **UniFrac (Weighted/Unweighted):** Phylogenetic; accounts for evolutionary distances between taxa.

> **Note:** For Aitchison, we use the CLR-imputed objects. For Bray, Jaccard, and UniFrac, we use Total Sum Scaling (TSS) to ensure sequencing depth doesn't bias the distances.

```{r plot different beta diversity measures, fig.width=12, fig.height=8}
# --- 1. CONFIG INPUT ---
beta_metric      <- config$Beta_Diversity$basic$metric
clr_variant      <- config$Beta_Diversity$basic$clr_variant
color_var        <- config$Beta_Diversity$color_var
shape_var        <- config$Beta_Diversity$shape_var
group_clustering <- config$Beta_Diversity$group_clustering

# --- 2. EXECUTION ---
source(config$scripts$beta_calc)
source(config$scripts$beta_vis_basic)

plot(p_beta)
```

## 5.2 Advanced Ordination: Environmental Correlation PCA variant

#In this section, we upgrade the ordination by mapping environmental variables (vectors) and group shapes (hulls) onto the microbial landscape.

```{r plot beta diversity measures PCA using aitchison distance, fig.width=12, fig.height=8}
# --- USER INPUT: Beta Diversity Configuration ---
# --- 1. CONFIG INPUT ---
beta_metric      <- "aitchison" 
clr_variant      <- config$Beta_Diversity$basic$clr_variant
color_var        <- config$Beta_Diversity$color_var
shape_var        <- config$Beta_Diversity$shape_var
group_clustering <- config$Beta_Diversity$group_clustering

# Advanced parameters
env_variables        <- config$Beta_Diversity$advanced$env_vars
top_asv_n            <- config$Beta_Diversity$advanced$top_asv_n
target_level         <- config$Beta_Diversity$advanced$target_tax_level
arrow_reduction_taxa <- config$Beta_Diversity$advanced$pca_arrow_taxa
arrow_reduction_env  <- config$Beta_Diversity$advanced$pca_arrow_env

# --- 2. EXECUTION ---
source(config$scripts$beta_calc)
source(config$scripts$beta_vis_pca)

plot(p_beta_PCA)
```

## 5.3 Advanced Ordination: Environmental Correlation PCoA variant

```{r plot beta diversity measures PCoA using various distance matrices as input, fig.width=12, fig.height=8}
# --- USER INPUT: Beta Diversity Configuration ---
# --- 1. CONFIG INPUT ---
beta_metric      <- config$Beta_Diversity$basic$metric # Usually wunifrac/bray
color_var        <- config$Beta_Diversity$color_var
shape_var        <- config$Beta_Diversity$shape_var
group_clustering <- config$Beta_Diversity$group_clustering

# Advanced parameters (PCoA specific scaling)
env_variables        <- config$Beta_Diversity$advanced$env_vars
top_asv_n            <- config$Beta_Diversity$advanced$top_asv_n
target_level         <- "class" # Or pull from config
arrow_reduction_taxa <- config$Beta_Diversity$advanced$pcoa_arrow_taxa
arrow_reduction_env  <- config$Beta_Diversity$advanced$pcoa_arrow_env

# --- 2. EXECUTION ---
source(config$scripts$beta_calc)
source(config$scripts$beta_vis_pcoa)

plot(p_beta_PCoA)
```

## 6. Differential_abundance_LinDA.R
### We use the filtered and not normalised phyloseq object, since LinDA handels the normalisation
### To make plots readable agglomerate at the selected phylogenetic level
### select the phylogenetic level to what the taxa should be agglomerated.
In this script we do a differential abundance analysis. In the following chunk calculations are made.
```{r, input script for LinDA here the model is defined and differental abundance calculations are done}
# --- 1. PREPARATION ---
target_level_linda <- config$Differential_Abundance$target_level
source(config$scripts$linda_prepare)

# --- 2. MODEL SETUP ---
l_mod <- config$Differential_Abundance$model

# Ensure the formula is a single string (R sometimes imports YAML as a list)
linda_formula_string <- paste(l_mod$formula, collapse = "")

# Set Reference dynamically
meta_df[[l_mod$main_var]] <- relevel(factor(meta_df[[l_mod$main_var]]), 
                                     ref = l_mod$ref_group)

# --- 3. RUN LINDA ---
l_params <- config$Differential_Abundance$params
linda_res <- linda(
  feature.dat     = otu_tab, 
  meta.dat        = meta_df, 
  formula         = linda_formula_string,
  feature.dat.type = "count",
  prev.filter     = l_params$prev_filter,
  is.winsor       = TRUE,
  outlier.pct     = l_params$outlier_pct,
  zero.handling   = l_params$zero_handling,
  pseudo.cnt      = l_params$pseudo_cnt,
  alpha           = l_params$alpha,
  p.adj.method    = l_params$p_adj_method
)

# From this linda output you can plot multiple things
```

# Plotting the Linda results
The interpretation of these results must be done carefully depending on the Linda input, Think about the model formula, and reference level since these will be essential in interpreting the output

For example, if you are working with a model which includes an interaction term. If you specify the reference group as above groundwater, then if you plot the 'SUM.PAHs..EPA.' then you will get how the bacteria differ in log fold change as if the conditions were above groundwater. 

Now if we are interested in how the bacteria differ in log fold changes with respect to PAH level below groundwater level, we need to change the reference group to below groundwater level to base the model on.

There is the possibility: "Groundwater_levelbelow:SUM.PAHs..EPA." which will indicate how the taxa 'respond to PAH levels below groundwater level' but the interaction term changes the interpretation greatly. So be carefull with the interpretation.
```{r plot LinDA results, fig.width=config$Differential_Abundance$plotting$fig_width, fig.height=config$Differential_Abundance$plotting$fig_height}
# Define which effect to visualize from Config
var_to_plot <- config$Differential_Abundance$plotting$var_to_plot

# Run the plotting script
source(config$scripts$linda_plot)
```
